---
title: "Math 248: Exploration PB"
author: "Your name here "
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(ImSIwithR)
library(tidyverse)
library(broom)
```


## Exploration PB: Housing Prices in Michigan

# STEP 1: Ask a research  question.

> 1. I want to predict how much should I expect to pay for a home. What explanatory variables might explain variation in home prices?.


# STEP 2: Design a study an collect data.

In 2015, we sampled 13 homes that were for sale just north of Lake Macatawa in Michigan, and recorded the sale price (in thousands of dollars) as listed on Zillow.

> 2. Identify the observational units and the response variable of interest. Is the response variable quantitative or categorical? What kinds of graphs and numerical summaries can you use to explore these data?


# STEP 3: Explore the data.

Since `price` is a quantitative variable, histogram is one of the appropriate plots to visualize this variable.
First we need to `read` data into R.

The data in `housing_data.csv` file posted on Canvas. Upload the `housing_data.csv` file to R studio.

Then, read this `housing_data.csv` file into R Markdown with the commands included in the R chunk below.

*Hint*: to get the correct path to include between the quotation marks, locate the csv file in the *Files* tab, click on the down arrow to the right of *More(Blue Wheel Icon)* and choose *Copy folder path to Clipboard* from the drop-down menu.

In my case the path copied to Clipboard is *"\~/Documents/Math 248 Materials/Math 248 Resources/Data"*. Paste this path before `/housing_data.csv` within the quotation marks in R code chunk below. Thus, *my full path* is :*"\~/Documents/Math 248 Materials/Math 248 Resources/Data/housing_data.csv".*

Use `head` command to make sure that the file is read correctly.

Remove **eval=FALSE** after running the code.

```{r, read the data from CSV file }
library(readr)
housing_data <- read_csv("~/Documents/Math 248 Materials/Math 248 Resources/Data/housing_data.csv")
#head(housing_data)
```

```{r,create a histogram of the price variable with ggplot}
# Basic histogram
ggplot(housing_data, aes(x = price)) +
  geom_histogram(binwidth = 150, fill = "lightblue", color = "black")
```


> 3. Describe the shape of the home prices in this sample.



> 4. Suppose we want to estimate or predict a typical home price based on these 13 homes. Do you think the mean would be a reasonable value to use? Explain why or why not.


> 5. Do you think the mean is larger or smaller than the median? Explain.



Let's check your answer:

```{r, numerical summary of the price variable}
summarize(housing_data,  
          mean_price = mean(price), 
          median_price = median(price),
          sd_price = sd(price, na.rm = TRUE)
          )
```


> 6. What is the standard deviation of these home prices from the output above? (What are the units?) Provide a one-sentence interpretation of this value. (Hint: How accurately can we predict a home price?) What are some possible explanations for this variation in home prices? (Of course, we also need to be a little wary in using the standard deviation with skewed data like this, as the deviations from the mean tend to be smaller for homes below the mean compared to homes above the mean.)



> 7. The first home in the data set sold for $639,000. If we had used the mean to predict that sale price, how far off would we have been?


> 8. Using the mean as our prediction, did we over-predict or under-predict the price of the first house. (Hint: Is the residual positive or negative?) 




Now, let's calculate the differences between each price value and the average price and name them `residual_mean`. We then create the histogram of the residuals from the mean to answer the question: 

> 9. Roughly how often do you find a positive residual (under-prediction) and how often do you find a negative residual (over-prediction)?.



```{r, residuals of the model that only uses mean price to predict price}
residuals_mean<-housing_data %>% 
  mutate(diff_from_mean = price - mean(price))

# Plot the histogram of the residuals ( differences from the mean)
ggplot(residuals_mean, aes(x = diff_from_mean)) +
  geom_histogram(binwidth = 150, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Residuals from the Mean",
       x = "Residual from the Mean Value",
       y = "Frequency") +
  theme_minimal()

```





_Let's also calculate the mean and the standard deviation of the distribution of the residuals from the mean._


```{r, Calculating the average and standard error of the residuals from the mean}
options(scipen = 999)#avoids scientific notation
summarize(residuals_mean,  
          mean_residuals = mean(diff_from_mean), 
          SE_residuals = sd(diff_from_mean, na.rm = TRUE)
          )
```
_In fact, when the mean is used to predict the observed response values, the average of the residuals will always be zero (within rounding)._


**Definition**

We will define a **statistical model** as consisting of the equation that predicts the outcome of the response and a measure of the precision of those predictions.


For example, here we could use the following statistical model:

$$Predicted \ \ price=408.062  \ \ thousand \ \ dollars$$

$$Standard \ \ Error \ \ of \ \ residuals = 240.92 \ \ thousand \\ dollars$$

_The **Standard Error (SE) of the residuals** is a measure of a typical prediction error from the model. Because the above model uses the overall mean as the prediction, the residual standard error is the same as the standard deviation of the home prices. Using more complex models will change the residual SE. We’ll discuss the computation of the residual SE more in Chapter 1._

The residual SE for this statistical model is still pretty large considering the values of these home prices (the standard deviation is about half the size of the mean). With so much unexplained variation remaining, this is a pretty uninformative model; can we improve our model using more information about these homes? 

_Let’s use information about the size of the house in our predictions to see whether we can reduce the sizes of the residuals overall. First, since both variables are quantitative we can create a scatterplot to visualize the relationship. `price` is the response variabe and should be on the y-axis,`sqft` is the explanatory variable that should be on the x-axis._

```{r, scatterplot of price vs size}
ggplot(data = housing_data, aes(x = sqft, y = price)) +
  geom_point()
```


> 10. Based on the scatterplot above describe the association between home price and square footage (strength, direction, linearity?). Does the association behave as you would have predicted? Explain.





Let's estimate a new `statistical model` predicting price using the size of the house (`sqft`). First, let's add the regression line to the scatterplot.

```{r, Scatterplot of price vs size with added regression line}
ggplot(data = housing_data, aes(x = sqft, y = price)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Scatterplot of Price vs. Size",
    x = "Size of the house in square ft",
    y = "Price of the house"
  ) 
```


Now, let's obtain the `estimated equation of regression line`:

```{r, Fit the linear model using sqfr to predict price}

model_fit <- lm(price ~ sqft, data = housing_data)

# Get a tidy tibble of the coefficients and statistics
tidy_output <- tidy(model_fit)
print(tidy_output)
```

Thus the predicted/fitted regression line is:

$$\widehat{Price}=-59.3687 + 0.2127 sqft$$



> 11.  Using the statistical model:

>> a. Provide interpretations, in context, of the slope and intercept of the line.

>> b. What residual does this give you for the first home in the dataset? Is this residual smaller or larger than before?

```{r, Add residuals and fitted values back to the original data}
augmented_housing_data <- augment(model_fit, newdata = housing_data)
head(augmented_housing_data, n = 5)
```

>> c. From the R output below, is the standard error of the residuals larger or smaller than before (it was around 240)? Explain what this tells you.

```{r, Rename the residuals, calculate mean and SE of residuals using sqft to predict price}
augmented_housing_data <- augmented_housing_data  %>% 
  mutate(regression_residuals = .resid) 


options(scipen = 999)# avoids scientific notation
summarize(augmented_housing_data,  
          mean_regression_residuals = mean(regression_residuals), 
          SE_regression_residuals = sd(regression_residuals, na.rm = TRUE)
          )

```
_Now consider the location of the home. In this dataset, the homes have been classified as lake front or not. Let's use R to examine separate regression lines for lake front and non-lake front homes._


```{r, scatterplots of price vs sqft for each lake location}
ggplot(housing_data, aes(x = price, y = sqft, color = lake)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # 'se=FALSE' removes confidence bands
  labs(title = "Price vs Size by Lake", x = "Price of the house", y = "Size of the house in sqft")
```


> 12)  Let's Fit separate regression lines for lakefront and non-lakefront homes. How do the two lines compare (think about both the slopes and the intercepts). Do these comparisons make sense in this context?


```{r, Model for lake front}
# Create separate models for each 'lake' level
model_lake1 <- lm(price ~ sqft, data = housing_data[housing_data$lake == "lakefront", ])
# Get tidy output 
tidy_lake1 <- tidy(model_lake1)
print(tidy_lake1)
```

```{r, Model for not lake front}
model_lake2 <- lm(price ~ sqft, data = housing_data[housing_data$lake == "notlakefront", ])
tidy_lake2 <- tidy(model_lake2)
print(tidy_lake2)
```
> 13) Use this model to predict the price of the first home in the data set. (Hint: the first house is lakefront)

```{r, predicting the price of the first house in the dataset using lake model}
Lake_housing_data <- augment(model_lake1, newdata = housing_data) %>% filter(lake=="lakefront")
head(Lake_housing_data, n=1)
```


> 14) Find the Standard Error of the residuals for this model

```{r, mean and SE of the residuals of the lake model}
model_interaction <- lm(price ~ sqft * lake, data = housing_data)

Lakes_housing_data <- augment(model_interaction, newdata = housing_data) %>%
                      mutate(lakes_regression_residuals = .resid) 

options(scipen = 999)# avoids scientific notation
summarize(Lakes_housing_data,  
          mean_lakes_regression_residuals = mean(lakes_regression_residuals), 
          SE_lakes_regression_residuals = sd(lakes_regression_residuals, na.rm = TRUE)
          )
```

> 15) Does including this additional location variable further reduce the standard error of the residuals?


> 16) Based on the scatterplot, is location (lake front or not) confounded with size of the home? Remember to consider both whether location appears to be related to price and whether location appears to be related to size. (Include any additional output you use to help you answer this question.)


## STEP 4: Draw inferences beyond the data.

In later chapters, we will discuss how to use confidence intervals and prediction intervals to improve statements of our predictions, their accuracy, and the reliability of the methods.

# STEP 5: Formulate conclusions.

> 17) Summarize the conclusions you would draw from this study, including which statistical model you would recommend and why. Is there a larger population of homes to which you are willing to generalize your conclusions? Is it reasonable to conclude that either the size of the home and/or the location of the property is causing variation in home prices? Explain.

# STEP 6: Looking back and ahead.

> 18) Identify any limitations you see to this study. What additional data would you like to collect to answer this research question?

