---
title: "Math 248: Explorations 1.4-1.5"
author: "Your name here"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
library(ImSIwithR)
```

# Exploration 1.4: Golden Squirrels

Bergmann’s rule (named after German biologist Carl Bergmann) is an ecogeological rule that states that within a species specimen will tend to be larger if they are from cooler climates or more extreme latitudes. Bergmann’s rule is most often applied to warm-blooded animals, but there has been some evidence of the rule in other species as well. A former Cal Poly Biology grad student (Nora Gerdes) wanted to investigate whether Bergmann’s Rule applies to the golden mantled squirrel in California. She measured the *body lengths (mm)* of 18 squirrels from four California *locations*: Hemet, Big Bear Lake, Sasanville, and Loop Hill. The data are stored in `Squirrels` in the `ImSIwithR` library in R.

*First, let's load the data:*

```{r, loading squarrels data from the package}
library(ImSIwithR)# reference the library where the dataset is stored
data(Squirrels)#load in data
head(Squirrels)
```

# STEP 1: Ask a research question.

> 1.  State the research question along the lines of an alternative hypothesis. Conjecture any other possible sources of variation in squirrel length.

# STEP 2: Design a study and collect data.

> 2.  Identify the observational units, the response variable, and the explanatory variable. Classify each.

> 3.  Is this an experiment or a observational study? Include a discussion of any inclusion criteria.

> 4.  Fill in the list of the sources of Variation components for this study:

-   Observed Variation in:\
-   Inclusion criteria:
-   Sources of explained variation:
-   Sources of unexplained variation:

*Let's summarize the distribution of squirrel lengths, using both a graphical summary (histogram & boxplot) and summary statistics.*

```{r, numerical summary of the length variable}
summarize(Squirrels,  
          mean_length = mean(Length), 
          sd_length = sd(Length, na.rm = TRUE)
          )
```

```{r, histogram of length variable}
ggplot(Squirrels, aes(x = Length)) +
  geom_histogram(binwidth = 15, fill = "lightblue", color = "black")+
  labs(title = "Histogram of squirrel lengths",
       x = "Length",
       y = "Count") +
  theme_minimal()
```

> 5.  Describe what you see on the graphs and what the numerical summaries tell you about this data.

> 6.  What is the `SSTotal` for this data?

```{r, calculating SSTotal}
overall_mean_val <- mean(Squirrels$Length, na.rm = TRUE)
print(paste("overall_mean:", overall_mean_val))

SSTotal <- sum((Squirrels$Length - overall_mean_val)^2)
print(paste("SSTotal (calculated from raw):", SSTotal))
```

*Now look at the graphical and numerical summaries of lengths separated by location.*

```{r}
# Create the stacked histogram
ggplot(Squirrels, aes(x = Length, fill = Location)) + 
  geom_histogram(binwidth = 10, color = "white") + 
  facet_wrap(~ Location, ncol = 1, scales = "free_y") +
  labs(title = "Histograms of Length by Location (Stacked Vertically, Colored by Group)",
       x = "Length",
       y = "Count") +
  theme_minimal()
```

```{r, box-plots of time for each treatment}
# Create boxplots 
ggplot(Squirrels, aes(x = Length, y = Location, fill = Location)) +
  geom_boxplot(color = "black") + # 'color' sets the outline of the boxes
  labs(
    title = "Boxplots of Length by Location",
    x = "Location",
    y = "Length"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Hide legend if fill is mapped to x-axis already

```

```{r, means and standard deviations of the Location groups}
summary_data <- Squirrels %>%
  group_by(Location) %>%
  summarise(
    mean_length = mean(Length, na.rm = TRUE), # Calculate mean
    sd_time = sd(Length, na.rm = TRUE),       # Calculate standard deviation
    n = n(),                                  # Count the number of observations
   .groups = 'drop'                           # Drop the grouping structure
  )

# View the new summary data frame
print(summary_data)
```

> 7.  Does location appear to explain variation in squirrel lengths? How are you deciding?

> 8.  Use group means to write out a "seperate locations" model. Give the predicted equation and standard error of residuals. (Hint: replace `?` with appropriate values).

$$
predicted \ length = 
\begin{cases} 
? & \text{if ?} \\
? & \text{if ?} \\
? & \text{if ?} \\
? & \text{if ?}
\end{cases}
$$

```{r}

model_squirrels <- lm(Length ~ Location, data = Squirrels)
rse_value <- sigma(model_squirrels) # residual standard error
print(paste("Residual Standard Error:", rse_value))
```

$Standard \ Error \  of  \ residuals = ? \  mm$

## Practical Significance:

Recall that in evaluating practical significance, you should consider the context of the research study as well as a numerical measure of group differences. When we have only two groups, we can compare the difference in means to the residual SE: $\bar{y_1}-\bar{y_2}/(Residual\ SE)$. If the difference in means is larger than one residual SE,

We tend to think of the difference as practically significant. When there are more than two groups, we can use $R^2$ to summarize how different the groups are in terms of how much variation in the response variable we are able to explain.

> 9.  Are any pairs of group means farther apart than one residual standard error? In other words, are effect sizes large enough to conclude practical significance?

```{r, effect size calculation }
# compute means and check locations
observed_means <- Squirrels %>%
  group_by(Location) %>%
  summarise(mean_length = mean(Length, na.rm = TRUE)) %>%
  ungroup()

# Extract the Standard Error of the Residuals 
model_squirrels <- lm(Length ~ Location, data = Squirrels)
rse_value <- sigma(model_squirrels)

# create pairwise combinations and compute effect sizes
pairwise_effects <- observed_means %>%
  rename(mean = mean_length) %>%
  mutate(key = 1) %>%
  inner_join(., ., by = "key", suffix = c("_A", "_B")) %>%
  filter(Location_A < Location_B) %>%
  transmute(
    pair = paste(Location_A, "vs", Location_B),
    mean_diff = mean_A - mean_B,
    effect_size = mean_diff / rse_value
  )

print(pairwise_effects)  # shows each pair, difference, and effect size

```

> 10. Report and Interpret $R^2$ for this data.

```{r, extracting the r-squared value from lm function output}
# 1. Fit the model
model_squirrels <- lm(Length ~ Location, data = Squirrels)

# 2. Get the R-squared value
observed_r2<-summary(model_squirrels)$r.squared
cat("Observed R-squared:", round(observed_r2, 4), "\n")
```

# STEP 4: Draw inferences beyond the data. (Statistical Significance)

*Practical significance* is only part of the story. We also want to consider whether the observed differences in mean lengths among the four locations are plausibly due to random chance alone or strong evidence of a true association between location and length.

> 10. Write out in words the null and alternative hypotheses for our significance test. (Hint: You may do so simply in terms of “association” between the response and explanatory variables or in terms of population means - being sure to define any symbols that you use.)

## Applying the 3S Strategy

1.  *Choice of Statistic*

First, we need one number that summarizes how different the groups are. The difference in means and t-statistic only work when we have just two groups to compare.

> 11. Suggest a statistic (a formula) that we could use to summarize the differences among these four groups.

2.  *Simulation*

There are several different statistics we could use; the key is getting some sense overall of how different the squirrel lengths are among the four location groups using *only one number*. This study was not a randomized experiment, but we can still shuffle the observed lengths to the four locations many, many times and determine how often we would randomly get a value for this statistic as or more extreme than that found in the actual environments. So, after each shuffle, as before, we will need to calculate the value of this statistic, and build a null distribution for the statistic. As we’ve already computed $R^2$, let’s start with $R^2$ as our statistic.

*Let's create the null distribution of the* $R^2$ statistic

```{r, creating the null distribution of r-squared statistic, echo=FALSE}
# --- Step 1: Set a seed for reproducibility ---

set.seed(123) 

# --- Step 2: Define the number of simulations ---
num_simulations <- 1000
simulated_r2s <- numeric(num_simulations) # Variable to store R-squared values

# --- Step 3: Perform the randomization (shuffling) test ---
for (i in 1:num_simulations) {
  # Shuffle the 'time' variable
  shuffled_lengths <- sample(Squirrels$Length)

  # Create temporary dataframe
  temp_data <- data.frame(
    length = shuffled_lengths,
    location = Squirrels$Location
  )

  # Fit a linear model to the shuffled data
  sim_model <- lm(length ~ location, data = temp_data)
  
  # Extract the R-squared value
  simulated_r2s[i] <- summary(sim_model)$r.squared
}

# --- Step 4: Visualize ---
sim_data_for_plot <- data.frame(r_squared = simulated_r2s)

ggplot(sim_data_for_plot, aes(x = r_squared)) +
  geom_histogram(binwidth = 0.02, fill = "skyblue", color = "black", alpha = 0.7) +
  
  # Vertical line for observed R-squared
  geom_vline(xintercept = observed_r2, color = "red", linetype = "dashed", size = 1) +
  
  labs(
    title = "Null Distribution of R-squared",
    subtitle = "What R-squared values would we see if Location had NO effect?",
    x = "Simulated R-squared",
    y = "Frequency"
  ) +
  theme_minimal() +
  
  # Annotations
  annotate("text", x = observed_r2, y = Inf, 
           label = paste("Observed R2 =", round(observed_r2, 3)), 
           vjust = 2, hjust = 1.1, color = "red", fontface = "bold") +
    # Expand Y axis to fit labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

```

> 12. Describe the behavior of the null distribution of the $R^2$ statistic. Is it roughly symmetric? Does this shape make sense? Explain.

3.  *Strength of Evidence*

*Let's use R to estimate the p-value.*

```{r, calculate P-value using r-squared distribution}
# NOTE: R-squared is always positive. A "stronger" result is always a HIGHER R-squared.
# Therefore, we only look at the right tail (greater than or equal to).
p_value <- sum(simulated_r2s >= observed_r2) / num_simulations
cat("P-value:", p_value, "\n")
```

> 13. Report the p-value. Does this p-value provide strong evidence against the null hypothesis of no association between location and length? Explain.

*Now, let's copy our data several times to create 4 copies of the original data.*

```{r}
copies <- rep(list(Squirrels), 4)
names(copies) <- paste0("Squirrels_copy_", 1:4)

four_squirrels_copies <- bind_rows(copies, .id = "copy")
head(four_squirrels_copies, 3)
```

> 14. How have summary statistics (means and standard deviations and sample sizes) changed?

```{r, means and standard deviations of the Location groups from 4 copies of the data}
summary_data_4 <- four_squirrels_copies %>%
  group_by(Location) %>%
  summarise(
    mean_length = mean(Length, na.rm = TRUE), # Calculate mean
    sd_time = sd(Length, na.rm = TRUE),       # Calculate standard deviation
    n = n(),                                  # Count the number of observations
   .groups = 'drop'                           # Drop the grouping structure
  )

# View the new summary data frame
print(summary_data_4)
```

> 15. How has the $R^2$ value changed?

```{r, extracting the r-squared value from lm function output for 4 copies of the data}
# 1. Fit the model
model_squirrels4 <- lm(Length ~ Location, data = four_squirrels_copies)

# 2. Get the R-squared value
observed_r2<-summary(model_squirrels4)$r.squared
cat("Observed R-squared:", round(observed_r2, 4), "\n")
```

> 16. Now reshuffle 1000 times. What changed in the null distribution?

```{r, creating the null distribution of r-squared statistic four copies of the data, echo=FALSE}
# --- Step 1: Set a seed for reproducibility ---

set.seed(123) 

# --- Step 2: Define the number of simulations ---
num_simulations <- 1000
simulated_r2s <- numeric(num_simulations) # Variable to store R-squared values

# --- Step 3: Perform the randomization (shuffling) test ---
for (i in 1:num_simulations) {
  # Shuffle the 'time' variable
  shuffled_lengths <- sample(four_squirrels_copies$Length)

  # Create temporary dataframe
  temp_data <- data.frame(
    length = shuffled_lengths,
    location = four_squirrels_copies$Location
  )

  # Fit a linear model to the shuffled data
  sim_model <- lm(length ~ location, data = temp_data)
  
  # Extract the R-squared value
  simulated_r2s[i] <- summary(sim_model)$r.squared
}

# --- Step 4: Visualize ---
sim_data_for_plot <- data.frame(r_squared = simulated_r2s)

ggplot(sim_data_for_plot, aes(x = r_squared)) +
  geom_histogram(binwidth = 0.02, fill = "skyblue", color = "black", alpha = 0.7) +
  
  # Vertical line for observed R-squared
  geom_vline(xintercept = observed_r2, color = "red", linetype = "dashed", size = 1) +
  
  labs(
    title = "Null Distribution of R-squared for 4 copies of the data",
    subtitle = "What R-squared values would we see if Location had NO effect?",
    x = "Simulated R-squared",
    y = "Frequency"
  ) +
  theme_minimal() +
  
  # Annotations
  annotate("text", x = observed_r2, y = Inf, 
           label = paste("Observed R2 =", round(observed_r2, 3)), 
           vjust = 2, hjust = 1.1, color = "red", fontface = "bold") +
    # Expand Y axis to fit labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

```

> 17. How does the p-value (our strength of evidence against the null hypothesis) change? Explain why this makes sense.

```{r, calculate P-value using r-squared distribution four copies of the data}
# NOTE: R-squared is always positive. A "stronger" result is always a HIGHER R-squared.
# Therefore, we only look at the right tail (greater than or equal to).
p_value <- sum(simulated_r2s >= observed_r2) / num_simulations
cat("P-value:", p_value, "\n")
```

*When we have more data, with the same differences in means and the same variability, we might find the additional “consistency” in the group differences more convincing, less likely to be due to “random chance” alone. The simulation reflects this, giving us a smaller p-value, but it would be nice if our statistic did so as well. The* $R^2$ statistic was the same whether we had 18 total observations or 72 total observations. Because the $R^2$ does not take sample size into account, we may prefer to look at a standardized statistic (something analogous to the t-statistic) that reflects both the sample size and the leftover or unexplained variation.

## Other Choices of Statistics The F-statistic is one such statistic.

Named after famous statistician R. A. Fisher, the F-statistic compares the explained to unexplained variance, adjusting for the sample size and number of groups, using the degrees of freedom for both the `SSError` and `SSModel`.

$$F= \left[ \frac{R^2}{1-R^2}\right]\times\left[\frac{n-\# \ of \ groups}{\#\ of \ groups - 1}\right]= \left[\frac{SSModel}{SSError}\right]$$

*Let's create the null distribution for the F-statistic using the four copies of the data.*

```{r, null distribution for the F-statistic }
# compute observed F
obs_aov <- aov(Length ~ Location, data = four_squirrels_copies)
obs_F <- summary(obs_aov)[[1]]["Location", "F value"]

# permutation function
perm_F <- function(df) {
  df_perm <- df %>% mutate(Location = sample(Location))  # shuffle Location labels
  sumsq <- summary(aov(Length ~ Location, data = df_perm))[[1]]
  sumsq["Location", "F value"]
}

# build null distribution
set.seed(123)                # for reproducibility
n_perm <- 2000
null_F <- replicate(n_perm, perm_F(four_squirrels_copies))

# p-value (one-sided: proportion of null F >= observed F)
p_val <- mean(null_F >= obs_F)

# optional: view null summary or plot
#summary(null_F)
hist(null_F, breaks = 40, main = "Null distribution of F", xlab = "F", col = "lightblue")
abline(v = obs_F, col = "red", lwd = 2)
```

```{r, observed F-statistic and the corresponding p-value}
cat("Observed F for 4 copies:", round(obs_F, 3), "\n")
cat("Permutation p-value for 4 copies:", signif(p_val, 3), "\n")
```

> 18. Report the observed F statistic and corresponding p-value.

> 19. Consider the original `Squirrels` dataset with 18 observations. How does the F-statistic for the original data compare to the F-statistic for the four copies dataset? As you expected? The p-value?

```{r, observed F statistic for original data}
obs_aov <- aov(Length ~ Location, data = Squirrels)
obs_F_org <- summary(obs_aov)[[1]]["Location", "F value"]
cat("Observed F for original squirrels data:", round(obs_F_org, 3), "\n")
```

```{r, p-value corresponding to F statistic for original data}
set.seed(123)                # for reproducibility
n_perm <- 2000
null_F_org <- replicate(n_perm, perm_F(Squirrels))

# p-value (one-sided: proportion of null F >= observed F)
p_val_org <- mean(null_F_org >= obs_F_org)
cat("Permutation p-value for original data:", signif(p_val_org, 3), "\n")
```

## Theoretical F Distribution:

One advantage of the *F-statistic* is when certain *validity conditions are met*, it is well-approximated by a probability distribution, the *F-distribution* (also named after R. A. Fisher).

### Validity Conditions:

To use the *F-distribution* to find the p-value for the F-statistic requires that

1.  the samples are independent of each other,
2.  the standard deviations of the treatment groups are similar (e.g., the largest is not more than twice the size of the smallest), and
3.  the distributions of the treatment groups are approximately symmetric (implying the distribution of the residuals is approximately normal) or all group sizes are larger than 20 with no extreme skewness or outliers.

Notice that these *validity conditions* are the same conditions we used for the two-sample (pooled) t-test.

> 20. Consider the original dataset with 18 observations. Go through each validity condition and determine whether or not they are met for this study. Hint: Use description of the data for independence, use standard deviations for each location and boxplots created in the beginning of the exploration to justify your answers.

> 21. Let's overlay F-distribution on the simulated null F distribution. Does the theoretical F-distribution do a good job of approximating the shuffled null distribution even in the original study with such small sample sizes?

```{r, overlaying theoretical F distribution}
# degrees of freedom
k <- length(unique(Squirrels$Location))
n <- nrow(Squirrels)
df1 <- k - 1
df2 <- n - k

# plot histogram as density, simulated density, theoretical F, and observed F
hist(null_F_org, breaks = 40, prob = TRUE, col = "lightblue",
     main = "Null distribution of F with theoretical F overlay",
     xlab = "F statistic")
lines(density(null_F_org), col = "blue", lwd = 2) 
# theoretical F density: choose reasonable x range covering the histogram
x_min <- min(null_F_org)
x_max <- max(null_F_org)
curve(df(x, df1, df2), from = x_min, to = x_max, add = TRUE, col = "red", lwd = 2)

abline(v = obs_F_org, col = "black", lwd = 2, lty = 2)
legend("topright", legend = c("Simulated density", "Theoretical F", "Observed F"),
       col = c("blue", "red", "black"), lwd = c(2,2,2), lty = c(1,1,2))
```

> 22. Based on the permutation p-value for the original data, what statistical decision can you make regarding the null hypothesis?

*Note:* the permutation/simulated and theory-based p-values should be similar since F distribution and shuffled distributions are similar. You could obtain the theory-based p-value using the following function in R.

```{r, theory-based p-value for ANOVA}
model_squirrels <- lm(Length ~ Location, data = Squirrels)
ANOVA_table <- anova(model_squirrels)
p_value <- ANOVA_table["Location", "Pr(>F)"]
cat("theory-based p-value for ANOVA:", round(p_value, 4), "\n")
```

# STEP 5: Formulate conclusions.

> 23. Based on your analysis so far, summarize the conclusions you would draw from this study. Be sure to address statistical significance, generalizability, and causation. Also be sure to put your comments into the context of this research study and your conclusions about Bergmann’s rule.

# STEP 6: Look back and ahead.

> 24. Suggest at least one way you would improve the study if you were to carry it out, or an extension to the research.

## Exploration 1.5 Golden Squirrels cont.

But, have we really answered the research question? Not yet! So far, we’ve only found evidence of an association between length and location. To determine whether Bergmann’s Rule applies, we need to understand the nature of the association between length and location. We need to address questions such as:

• Do squirrels from colder locations tend to be longer, on average?

• Which population mean or means is/are different than the others? How much do they differ?

• Does the average length differ in every one of these four populations? Maybe just one of them?

## Post-hoc Analyses

Once we find a significant association, the natural follow-up question is the nature of that association. The process of following up a *statistically significant F-test* is called *post-hoc analysis*.

## Definition

A *post hoc analysis* is the process of assessing how the means of the treatment groups relate to one another, a follow up analysis to a significant F-Test.

### Pairwise Comparisons of the Treatment Groups:

Arguably the most common type of post-hoc analysis involves comparing each group mean to each other group mean by conducting *pairwise comparisons*.

## Definition

The *experiment-wise Type I error rate* is the chance of making at least one Type I error when conducting numerous tests of statistical significance. To better control this rate, we will only conduct post-hoc analyses after a significant F-test.

## Key Idea

We can protect against an inflated experiment-wise Type I error rate by only conducting post-hoc analyses using pairwise comparisons after obtaining a statistically significant F-statistic.

*Let's use R to calculation 95% confidence intervals for the difference in means.*

```{r pairwise confidence intervals without adjustment}
# ANOVA and pooled residual SD
model <- aov(Length ~ Location, data = Squirrels)
aov_tab <- anova(model)
mse <- aov_tab["Residuals", "Mean Sq"]        # mean squared error
sigma <- sqrt(mse)
# group means and counts
gm <- Squirrels %>%
  group_by(Location) %>%
  summarise(mean = mean(Length, na.rm = TRUE), n = n()) %>%
  arrange(Location)

# critical t for (1-alpha/2)
df_res <- df.residual(model)
tcrit <- qt(0.975, df_res)   # 95% CI

# all pairwise combinations and CIs
pairs <- expand.grid(i = 1:nrow(gm), j = 1:nrow(gm)) %>%
  filter(i < j) %>%
  rowwise() %>%
  mutate(
    loc_i = gm$Location[i],
    loc_j = gm$Location[j],
    mean_diff = gm$mean[i] - gm$mean[j],
    se_diff = sigma * sqrt(1 / gm$n[i] + 1 / gm$n[j]),
    lower = mean_diff - tcrit * se_diff,
    upper = mean_diff + tcrit * se_diff
  ) %>%
  select(loc_i, loc_j, mean_diff, se_diff, lower, upper)

pairs
```

This output mathes the output produced by the `Comparing Groups`applet.

> 1.  Do any of these confidence intervals contain 0? If so, which one(s) and what does it mean when the confidence interval for a difference of means contains 0?

*Since these are created after determining that the F-test produced a statistically significant result, these are called Fisher's Least Squared Differences. Fisher's Least Significant Difference (LSD) produces unadjusted pairwise confidence intervals for comparing group means in ANOVA. Because it treats each comparison independently rather than adjusting for the total number of comparisons, it is more likely to find significant differences (high power) but at the cost of higher Type I error rates. It is generally used when the ANOVA F-test is significant, particularly with a small number of pairs.*

## Tukey’s Honest Significant Difference (Tukey’s HSD) method

A commonly used method to make all the pairwise comparisons that includes a correction to adjust a Type I error rate is called *Tukey’s Honest Significant Difference* (Tukey’s HSD) method. Using this method allows simultaneous confidence: the 95% confidence applies to the entire set of comparisons, not just individual pairs.

*Let's use R to create Tukey-adjusted confidence intervals:*

```{r, Tukey adjusted confidence intervals}
# fit ANOVA
model <- aov(Length ~ Location, data = Squirrels)

# Tukey HSD gives differences and CIs
tuk <- TukeyHSD(model, "Location", conf.level = 0.95)
tuk
# as a data.frame
as.data.frame(tuk$Location)
```

> 2.  Do any of these Tukey - adjusted confidence intervals contain 0? How does the width of adjusted confidence interval compare to unadjusted Fisher LSD confidence intervals? Did the significance of the pairwise comparisons changed by using the adjusted confidence interval?

> 3.  Write a brief summary of your findings from the pairwise confidence intervals. Be sure to address whether/how these support the application of Bergmann’s Rule to the golden mantled squirrel in California.

## Other multiplicity adjustments

There are several common *multiplicity adjustments*; some are more conservative (increase the p-value a lot) than others:

-   Fisher LSD (Least Significant Difference)

*What it is:* A two-step approach that first checks for any overall difference; if that test is significant, it then performs pairwise comparisons without further strong adjustment.

*When to use:* When you have a small number of groups and you’re willing to accept a higher chance of false positives for follow-up pairwise tests. Best when the overall test is significant and you want more power to detect differences.

*Trade-off:* More powerful but risks more false positives when many comparisons are made.

-   Tukey (Tukey’s Honest Significant Difference)

*What it is:* A method designed specifically for all pairwise comparisons that controls the overall error rate across every pair.

*When to use:* When you plan to compare every group against every other group (all pairwise contrasts). It balances protection against false positives with reasonable power and provides interpretable simultaneous confidence intervals.

*Trade-off:* Slightly more conservative than LSD (less chance of false positives) but generally a good default for multiple pairwise comparisons—recommended in most routine situations.

-   Scheffé

*What it is:* A very flexible, conservative method that protects against false positives for any possible contrast (not just pairwise).

*When to use:* When you may test many or complex contrasts (including custom combinations) and need strong protection against false positives across all of them.

*Trade-off:* Quite conservative, so it has lower power to detect true differences for specific pairwise tests.

Bonferroni

*What it is:* A straightforward correction that divides your significance threshold by the number of comparisons.

*When to use:* When you have a small, clearly specified set of planned comparisons and want a simple, easy-to-explain adjustment.

*Trade-off:* Can be overly conservative when many comparisons are made, reducing power.

### Recommendation

For routine studies comparing every group to every other group, Tukey’s method is a strong default: it controls the family-wise error rate, gives simultaneous confidence intervals, and balances avoiding false positives with reasonable power. Use Fisher LSD only when you accept a higher risk of inflating Type I error to achieve greater power (after a significant overall test); use Scheffé for broad protection over arbitrary contrasts; and use Bonferroni for a small number of planned comparisons when simplicity is preferred.

## Confidence Intervals on Other Parameters

The confidence intervals for the difference in means allow us to compare the mean lengths of squirrels from two locations. However, what if we also wanted to estimate the average length of the population of squirrels from Big Bear Lake?

## Definition

A *one-sample t-interval for a population mean* is:

$$\bar{y_i}\pm (t^*_{df})\frac{residual \ SE}{\sqrt{n_i}}$$

Where $n_i$ is the sample size of the treatment group, and the $t^*$ multiplier is again approximately 2 for 95% confidence intervals.

> 4.  Use R output to report 95% confidence intervals for the mean length of all squirrels within each location. Write an interpretation of one of these intervals in the context of this study.

```{r, one-sample confidence intervals}
conf_level <- 0.95

# overall (pooled) SD of Length across all observations
sd_length <- sd(Squirrels$Length, na.rm = TRUE)
N_total <- sum(!is.na(Squirrels$Length))
df <- N_total - 1
tcrit <- qt((1 + conf_level) / 2, df)

one_sample_cis_length_only <- Squirrels %>%
  group_by(Location) %>%
  summarise(
    n = sum(!is.na(Length)),
    mean = mean(Length, na.rm = TRUE),
    se = sd_length / sqrt(n),                    # uses overall SD only
    lower = mean - tcrit * se,
    upper = mean + tcrit * se
  ) %>%
  ungroup()

#cat("Overall SD of Length:", sd_length, " (df =", df, ")\n")
print(one_sample_cis_length_only)
```

> 5.  How would increasing the sample size of the treatment groups change these intervals? What if the confidence was changed to 99%?

> 6.  Will any of the computed confidence intervals allow you to predict the length of an individual new squirrel at a particular location? Why or why not?

## Prediction Intervals

Up until now, our focus has been on confidence intervals for population means (or differences in means). These intervals provide estimates of ranges of plausible values for the unknown population mean value. These confidence intervals for the mean do not allow us to make predictions about the lengths of individual squirrels (e.g., how long we would estimate a squirrel to be if we randomly sampled one more squirrel from a particular location).

### Definition

A *prediction interval* gives an interval of values within which we predict the response of a new individual observation (e.g., person) to occur with some degree of confidence. For example, a 95% *prediction interval* means we are 95% confident that the responses for 95% of individuals in the population will be captured in the interval.

$$\bar{y_i}\pm t^*\times(SE\ of\ residuals)\sqrt{1+\frac{1}{n}}$$

*Let's create prediction intervals*

```{r, one-sample prediction intervals}
conf_level <- 0.95

# overall SD of Length (used as residual_SE) and df for t*
sd_length <- sd(Squirrels$Length, na.rm = TRUE)
N_total <- sum(!is.na(Squirrels$Length))
df <- N_total - 1
tcrit <- qt((1 + conf_level) / 2, df)

prediction_intervals <- Squirrels %>%
  group_by(Location) %>%
  summarise(
    n = sum(!is.na(Length)),
    mean = mean(Length, na.rm = TRUE)
  ) %>%
  mutate(
    margin = tcrit * sd_length * sqrt(1 + 1 / n),
    lower_pred = mean - margin,
    upper_pred = mean + margin
  ) %>%
  ungroup()

print(prediction_intervals)
```

> 7.  Write an interpretation of one of the prediction intervals. Comment on how your interpretation of this interval differs from the interpretation of the confidence intervals.

*Note: If we were to use 2 as a multiplier instead of tcrit value that depends on degrees of freedom, we would obtain slightly less precise intervals*:

```{r}
conf_level <- 0.95  # kept for reference but not used for multiplier
multiplier <- 2     # use 2 instead of tcrit

# overall SD of Length (used as residual_SE)
sd_length <- sd(Squirrels$Length, na.rm = TRUE)
N_total <- sum(!is.na(Squirrels$Length))
df <- N_total - 1

prediction_intervals <- Squirrels %>%
  group_by(Location) %>%
  summarise(
    n = sum(!is.na(Length)),
    mean = mean(Length, na.rm = TRUE)
  ) %>%
  mutate(
    margin = multiplier * sd_length * sqrt(1 + 1 / n),
    lower_pred = mean - margin,
    upper_pred = mean + margin
  ) %>%
  ungroup()

print(prediction_intervals)
```

> 8.  Will increasing the sample size within the treatment groups have a large impact on the width of these prediction intervals? Explain why or why not.

> 9.  In general, for a group of interest, which is wider: a 95% prediction interval or a 95% confidence interval? Explain.

*As we’ve seen before, these theory-based intervals have certain conditions that must be met in order to be valid.*

## Validity Condition for Confidence Intervals and Prediction Intervals on Means

• The data distribution should be reasonably bell-shaped and symmetric, especially if the sample sizes are small. This condition is particularly important for prediction intervals. (With confidence intervals, the distribution of the sample mean should become more normal when the sample size increases, but the distribution of the responses themselves does not change shape as we increase sample size.)

• Note: For comparing a subset of two treatment group means, the confidence interval of the difference requires approximately equal standard deviations among all of the treatment groups. This is because these confidence intervals use the residual standard error from the separate-means model.

> 10. Based on the boxplots and standard deviations of the squirrel data, do you think that the validity conditions are met? How are you deciding? (Hint: you have checked the conditions for this data in question 20 of Exploration 1.4).
