---
title: "Math 248: Exploration 2.2"
author: "Your name here"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
library(ggplot2)
library(dplyr)
library(mosaic)
library(ImSIwithR)
```

```{r}
#load in data
data(Strawberries)
```

# Exploration 2.2 Strawberry Storage

A study (Smith & Skog, “Postharvest Carbon Dioxide Treatment Enhances Firmness of Several Cultivars of Strawberry,” American Society for Horticulture Science, 27(5), 1992) was conducted in order to investigate the effects of storage method on the firmness of strawberries. As strawberries are harvested, they are placed into storage containers called “clamshells.”

Because clamshells have holes in them, air can circulate from outside the clamshell into the clamshell and vice versa while the container still protects the fruit. To test the potential effect of changing the composition of the air in which strawberries are stored, researchers obtained 15 clamshells filled with strawberries of several different varieties. Five clamshells were randomly assigned to be stored in regular air (approximately 21% oxygen and 0.04% carbon dioxide) and five clamshells were randomly assigned to be stored in a modified air method (18% oxygen and 15% carbon dioxide). After two days of refrigeration at 0.5°C in one of the two different atmospheres, the strawberries were then left at room temperature in regular air for two more days. The firmness of the strawberries was then measured by the force (measured in newtons, N) needed for a probe to pierce the exterior of the strawberry. The remaining five clamshells were randomly assigned to a “control” condition which measured the strawberries immediately (not after four days like the other two conditions).



>1. Identify the observational units, the explanatory variable, and the response variable. Complete the hypothesized Sources of Variation table. 

 | Observed variation in | Sources of explained variation | Sources of explained variation |
 |-----------------------|--------------------------------|--------------------------------|
 | Inclusion criteria    |                                |                                |
 |                       |                                |                                |
 | Design                |                                |                                |



>2. Consider the following outputs: 

```{r}
ggplot(Strawberries, aes(Firmness))+
  geom_histogram(fill="red", color="white")+
  theme_classic()+
  labs(title = "Firmness Distribution")
```

```{r}
ggplot(Strawberries, aes(Firmness, fill = Storage))+
  geom_boxplot()+
  facet_wrap(~Storage, ncol=1)+
  theme_classic()+
  labs(title="Distribution of Firmness by Storage Type")
  
```


>a. Calculate the observed effects for each storage method: 

```{r}
# Calculate group means and overall mean
stats <- Strawberries %>%
  summarize(
    grand_mean = mean(Firmness),
    mean_Air = mean(Firmness[Storage == "Air"]),
    mean_Control = mean(Firmness[Storage == "Control"]),
    mean_Mod = mean(Firmness[Storage == "ModifiedAir"])
  )

# Calculate Effects (Group Mean - Grand Mean)
effects <- stats %>%
  transmute(
    Effect_Air = mean_Air - grand_mean,
    Effect_Control = mean_Control - grand_mean,
    Effect_Mod = mean_Mod - grand_mean
  )

print(effects)
```


>b. Does the storage method appear to explain some of the variability in the firmness levels? How are you deciding? 



>3, Consider the ANOVA table for the (one variable) separate means model and a simulated null distribution of the F-statistic.

```{r}
# Observed F-statistic for the One-Way ANOVA (ignoring variety)
model_1 <- aov(Firmness ~ Storage, data = Strawberries)
anova(model_1)

obs_F <- anova(model_1)$F[1]
```


> How much variation in firmness values is explained by the storage method? Based on the F-statistic and the null distribution, should we find a large p-value or a small p-value? How are you deciding? (Hint: Look back at 2.1, question 14.)




>4. Suggest some possible explanations for finding an insignificant result here.




## Blocking Decision: 

Rather than limiting themselves to one type of strawberry (i.e., using inclusion criteria of one type of strawberry) which would most likely have given them a smaller *SSError*, the researchers decided to include five different varieties of strawberries in the study allowing them to generalize to a more diverse population of strawberries. Suspecting that different varieties of strawberries have different firmness ratings, that is, believing variety is associated with firmness (the response variable), the researchers opted to block on strawberry variety. In this way the variation explained by variety will be accounted for in the analysis by removing it from the unexplained variation.



>5. Based on the following output, do the data confirm or refute the researchers’ prediction that different strawberry varieties tend to have different firmness ratings? Is there strawberry variety that tends to have very low firmness? Very high?

```{r}
# Visualizing Variety differences
ggplot(Strawberries, aes(x = Variety, y = Firmness, fill = Variety)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  theme_bw() +
  labs(title = "Firmness by Strawberry Variety")
```



>6. The following graph labels the observations in the graph in #5 by the storage method. For the Allstar variety, which storage method had the largest firmness? Is this the case for all of the varieties? Looking at this graph, do you think the storage method effect will be statistically significant after adjusting for variety? Explain your reasoning.


```{r}
# Plotting Variety and Storage on Firmness
ggplot(Strawberries, aes(x = Variety, y = Firmness, color = Storage)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "Firmness by Variety and Storage Method")
```


>7. Outline a study design for these 15 clamshells that will allow you to compare the 3 storage methods adjusting for the variation in firmness in the 5 varieties.


## Key Idea: RCBD (Randomized Complete Block Design)


To start, samples of harvested Allstar strawberries were randomly placed into three clamshells and each clamshell was then randomly assigned to one of the three storage methods. This procedure was implemented for the four remaining strawberry varieties, as well. This restricted random assignment guarantees that there is one clamshell of each variety for each storage method. See the diagram for an illustration of this random assignment process.


The experimental unit in this study is the clamshell, not the individual strawberries because the treatments were assigned to the clamshells.


A block study design creates blocks of experimental units that are similar to each other, randomly assigns the treatments within each block, and then analyzes the data in a way which accounts for block-to-block variations. When there are only two groups being compared a block study design is called a matched pairs design. The term block comes from the first block designs which were agricultural experiments in large fields where separate parts of the field were called “blocks.”


Modifying the Analysis to Consider Blocking Now that we know more about the actual way this study was designed, we know that an analysis that ignores this “restricted randomization” is incorrect. Instead, the analysis should reflect the study design that was used.



>8. Describe how you would simulate F-statistics in a way that “shuffles” the clamshells to the storage treatments, mirroring the actual study design, but assuming the null hypothesis is true.



# Analysis 1: Ignore Blocking Effect

>9, Record the mean and standard deviation of this “completely randomized” null distribution. So only the effect of the storage type is taken into account. 

Remove **eval=FALSE** after running the code.

```{r, eval = FALSE}
set.seed(1) #SET YOUR own seed. 

# Simulation 1: Completely Randomized (ignoring blocks)
null_distribution_CRD <- do(1000) * aov(Firmness ~ shuffle(Storage), data = Strawberries)

#filter for the F values
Fstat <- null_distribution_CRD%>%
  filter(!is.na(F))

#create a distribution
ggplot(Fstat, aes(x = F)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  labs(title = "Simulated Null Distribution of F-statistics, Ignoring Blocking Effect",
       x = "F-statistic",
       y = "Frequency") +
  theme_minimal()
```


Mean: 

Std Dev:



>10. Use the observed F-statistic and the simulated null distribution to approximate the p-value.

```{r}
# P-value for Analysis 1
prop(~ F >= obs_F, data = null_dist_1)
```






But we still want to know whether, after adjusting for variety, the variability explained by storage method is statistically significant. It appears likely, given that within each variety, the modified air method has consistently larger firmness values.




# Analysis 2: Revising the Simulation to Account for the Blocking

>11. Describe how the individual values changed for one of the strawberry varieties.



>12. Within the Allstar variety (black dots), did the storage method that each firmness value was associated with change? If so, how? (Shuffle Responses again if you need to watch the simulation again, to be sure what is happening.) Within each variety, how many clamshells are assigned to a storage method?



>13. What are the mean and standard deviation of the simulated null distribution of F-statistics shuffling within blocks? How do these compare to the mean and standard deviation obtained earlier using a completely randomized approach (Analysis #1)?


```{r}
set.seed(123)
# Simulation 2: Randomized Complete Block Design (Shuffling within blocks)
# We calculate the F-statistic for the model: Firmness ~ Storage + Variety
# But we only shuffle Storage within Variety
# Note: F-statistic of interest is for Storage

null_distribution_RCBD <- do(1000) * aov(Firmness ~ shuffle(Storage, groups= Variety)+ Variety, data = Strawberries)

#filter for the F values
Fstat_RCBD <- null_distribution_RCBD%>%
  filter(!is.na(F))

#create a distribution
ggplot(Fstat_RCBD, aes(x = F)) +
  geom_histogram(bins = 100, fill = "skyblue", color = "white") +
  labs(title = "Simulated Null Distribution of F-statistics for RCBD",
       x = "F-statistic",
       y = "Frequency") +
  theme_minimal()


```



>14. You should see less random variation in the simulated F-statistics when shuffling within each variety block (Analysis #2). Explain why this makes sense. That is, explain why the variation of the null distribution of F-statistics will tend to be smaller when we shuffle within blocks as compared to shuffling in a completely randomized way.


### Key Idea: 

The the null distribution is typically less variable when shuffling within blocks, which makes observed values of the statistic of interest more extreme, providing more evidence against the null hypothesis.




>15. How will the smaller standard deviation in the shuffled within blocks null distribution impact the strength of evidence provided by the observed F-statistic, 0.995 ≈ 1.0? Explain.




>16. How has the strength of evidence (p-value) against the null hypothesis changed when considering blocks (Analysis #2) vs. Analysis #1 which did not consider the blocks? Why?

```{r}
# Calculate Observed F for the Blocked Model
model_2 <- aov(Firmness ~ Storage + Variety, data = Strawberries)
obs_F_blocked <- anova(model_2)$F[1] # F-stat for Storage

# Simulated p-value using blocking
p_val_2 <- prop(~ F >= obs_F_blocked, data = null_dist_2)
p_val_2
```


Simulated p-value without blocking (Question #10):

Simulated p-value using blocking (just found in Question #16):






### Important Note:

We conducted multiple analyses (one that considered the blocking design and one that did not) on the same data in order to illustrate the benefits of blocking. In practice, we analyze data only in a manner that reflects the study design. Using multiple analysis methods on the same data did help to clarify for us just how important consideration of the blocking variable is in this case.


Because the actual study used blocking and we modified the simulation to account for the blocking variable, we need to modify the null and alternative hypothesis being tested as shown here.



**Null hypothesis** $H_0$: The long-run average firmness of strawberries in each of the three experimental conditions is the same, after accounting for differences in firmness between different varieties of strawberries.

**Alternative hypothesis**  $H_a$: At least one of the three experimental conditions have a different long-run average firmness of strawberries, after accounting for differences in firmness between different varieties of strawberries.


>17. Based on the prior modified null and alternative hypotheses and your simulation analysis considering the blocking variable, state a conclusion about the study.



>18. Calculate the block effect for each of the five varieties of strawberries by subtracting the overall average firmness (6.28) from the average firmness for each variety. Note: the block effects will be positive for varieties with above average firmness and negative for varieties with below average firmness.


```{r}

overall_mean <- mean(Strawberries$Firmness)

block_effects <- Strawberries %>%
  group_by(Variety) %>%
  summarize(
    variety_mean = mean(Firmness),
    block_effect = variety_mean - overall_mean
  )

print(block_effects)
```


>19. Now, compute “adjusted” firmness ratings for each clamshell by subtracting off the block effects.

```{r}
# Join effects back to main data and calculate adjusted firmness
Strawberries <- Strawberries %>%
  left_join(block_effects, by = "Variety") %>%
  mutate(Firmness_Adj = Firmness - block_effect)

head(Strawberries %>% select(Variety, Firmness, block_effect, Firmness_Adj))
```


>20. Compare the block-adjusted firmness values to the original firmness values. In terms of the variation, how do the two sets of data compare?

```{r}
sd(Strawberries$Firmness)
sd(Strawberries$Firmness_Adj)
```



>21. What impact will this reduction in the variation within the treatment groups have on the adjusted F-statistic?





### Key Idea: 

When simulating the null distribution of the F-statistic on the adjusted data you no longer need to re-randomize within the blocks—instead, re-randomize the treatments across all of the experimental units (ignoring the blocks). The blocks have already been accounted for in the analysis by adjusting the data up front. If strawberries had come to us with the same firmness (on the same scale), we wouldn’t need to re-randomize within the blocks.



>22. Compare the following null distribution to the one obtained earlier shuffling within blocks. How has the null distribution changed?

```{r}
# Analysis 3: Adjusted Data (Complete Randomization on Adjusted Data)
# We verify the validity of using adjusted data with a simple one-way ANOVA simulation
obs_F_adj <- anova(aov(Firmness_Adj ~ Storage, data = Strawberries))$F[1]

null_dist_adj <- do(1000) * anova(aov(Firmness_Adj ~ shuffle(Storage), data = Strawberries))



mean(null_dist_adj$F)
sd(null_dist_adj$F)
```




Mean:

Standard Deviation:


>23. What is the simulated p-value using the “adjusted” F-statistic from the adjusted data?


```{r}
prop(~ F >= obs_F_adj, data = null_dist_adj)
```



>24. In terms of the effect of storage method on firmness, what do you conclude after analyzing the block-adjusted F-statistic? How does this conclusion compare to the conclusion from Analysis #2, shuffling within blocks? Why does this make sense?



##: Validity conditions for ANOVA with a blocking variable

Recall that theory-based approaches like those you have been using require certain validity conditions to be met. Here are the validity conditions for ANOVA with a blocking variable.


### The F-distribution is a good approximation to the null distribution of the F-statistic as long as:


- Either the sample size is at least 20 for each of the groups being compared, without skewness or outliers; *or* if the sample sizes are less than 20 for each group, then the distribution of adjusted response variable within each group is approximately symmetric. 

- The standard deviations of the adjusted response variable samples in each group being compared are approximately equal to each other. (Largest SD is not more than twice the smallest SD)

- The experimental units are randomly assigned to the treatments within each block.


>25. Comment on how the analysis you just conducted meets the validity conditions by evaluating the three conditions listed in the Validity conditions box.



```{r}
#look for skewness and outliers or approximately symmetric
ggplot(Strawberries, aes(x = Storage, y = Firmness_Adj, fill= Storage)) +
  geom_boxplot() +
  labs(title = "Adjusted Firmness by Storage")
```

```{r}
#check Standard deviations
Strawberries %>%
  group_by(Storage) %>%
  summarize(sd_adj = sd(Firmness_Adj))
```




>26. Summarize the conclusions you would draw from this study, including generalizability and causation (for both the method and the variety).






