---
title: "Math 248: Exploration 1.6"
author: "Your name here "
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
#library(pwr)
#library(readr)
```

## Exploration 1.6 Who Is Spending More Time Parenting on Average?

In this exploration, we assume that you are familiar with the idea of *statistical power* as discussed in Example 1.6.

Suppose that you are planning a study to investigate whether female parents tend to spend more time (hours) parenting each week, as compared to male parents and will use a significance level of 5%. How many people should you plan to have in your study? Let’s investigate this question by conducting a *statistical power analysis*.

> 1.  Identify the explanatory variable and the response variable. Also, identify whether each is categorical or quantitative, as well as units of measurement for any quantitative variable(s).

> 2.  Explain why this would be an observational study rather than a randomized experiment.

> 3.  State the null and the alternative hypotheses for this study.

> 4.  Will the study use random sampling or random assignment, or both, to recruit the participants?

> 5.  What will be the Type I error rate in your proposed study? Explain, in context, what a Type I error would represent in your planned study?

> 6.  What size p-value will you need in order to find a statistically significant result?



_But, how large of a sample size do you need in each group (men and women) in order to get a convincing p-value? To investigate this question we will simulate random samples from different theoretical populations (rather than random shuffling) in R. We can explore how our study might play out if we were to run it._

Let’s assume that:

> (a) women parent 15 hours per week on average and that it is the same for men,
> (b) that the standard deviation of both male and female parenting distributions is 5 hours per week, and
> (c) that we decide to sample 10 male parents and 10 female parents for this study.

In the R code below choose your own `seed` for reproducibility.

```{r, simulates single random sample for women and men respectively}
# --- Parameters ---
n_per_group <- 10
mean_pop <- 15 # Same mean for both under the Null
sd_pop <- 5
num_simulations <- 1

# --- Simulation Function ---
simulate_diff_means <- function(n, mu1, mu2, sd) {
  sample1 <- rnorm(n, mean = mu1, sd = sd)
  sample2 <- rnorm(n, mean = mu2, sd = sd)
  return(mean(sample1) - mean(sample2))
}

# --- Step 1: Perform one simulation (like "Draw samples" in applet) ---
#set.seed(?) # Reproducibility
sample_women <- rnorm(n_per_group, mean = mean_pop, sd = sd_pop)
sample_men <- rnorm(n_per_group, mean = mean_pop, sd = sd_pop)

cat("--- Single Sample Summary ---\n")
cat("Women Mean:", round(mean(sample_women), 2), " SD:", round(sd(sample_women), 2), "\n")
cat("Men Mean:  ", round(mean(sample_men), 2), " SD:", round(sd(sample_men), 2), "\n")
cat("Difference in Means:", round(mean(sample_women) - mean(sample_men), 2), "\n")
```

> 7.  What mean did you get for the women? SD? What mean did you get for the men? SD? How different are the means?

> 8.  Which hypothesis are you assuming to be true in this simulation?

> 9.  Now, let's take 1000 samples. The graph shows the difference in the two sample means for each of the 1000 samples you took. At what value is this distribution centered? Why does that make sense? What is the standard deviation?

```{r, Null Distribution of Difference in Sample Means}
# --- Parameters ---
n_per_group <- 10
mean_pop <- 15 # Same mean for both under the Null
sd_pop <- 5
num_simulations <- 1000

# --- Simulation Function ---
simulate_diff_means <- function(n, mu1, mu2, sd) {
  sample1 <- rnorm(n, mean = mu1, sd = sd)
  sample2 <- rnorm(n, mean = mu2, sd = sd)
  return(mean(sample1) - mean(sample2))
}

# --- Step 1: Perform one simulation ---
set.seed(123) 
sample_women <- rnorm(n_per_group, mean = mean_pop, sd = sd_pop)
sample_men <- rnorm(n_per_group, mean = mean_pop, sd = sd_pop)

# --- Step 2: Perform 1000 Simulations (Null Distribution) ---
null_differences <- replicate(num_simulations, simulate_diff_means(n_per_group, mean_pop, mean_pop, sd_pop))

# --- Calculate Mean and Standard Deviation of the Null Distribution ---
mean_null_diffs <- mean(null_differences)
sd_null_diffs <- sd(null_differences)

# --- Step 3: Visualize ---
null_data <- data.frame(diff_means = null_differences)

ggplot(null_data, aes(x = diff_means)) +
  # 1. Base Histogram (Gray)
  geom_histogram(binwidth = 0.5, fill = "lightgray", color = "black") +
  
  # Add a vertical line for the mean of the null distribution
  geom_vline(xintercept = mean_null_diffs, color = "blue", linetype = "dashed", size = 1) +
  
  # Add annotations for mean and standard deviation
  annotate("text", x = mean_null_diffs, y = Inf,
           label = paste("Mean =", round(mean_null_diffs, 2)),
           vjust = 1, hjust = -1.5, color = "blue", fontface = "bold") +
  annotate("text", x = sd_null_diffs, y = Inf,
           label = paste("SD =", round(sd_null_diffs, 2)),
           vjust = 2.5, hjust = -0.5, color = "blue", fontface = "bold")+
   # 5. Labs and Theme
  labs(
    title = "Null Distribution of Difference in Sample Means (1000 simulations)",
    subtitle = "Assuming Null Hypothesis is True (Mean Diff = 0)",
    x = "Difference in Means (Women - Men)",
    y = "Count"
  ) +
  theme_minimal()
```

> 10. Use the null distribution created above to answer the following question: If you actually did this study (surveying 10 men and 10 women), and got a difference of 5 hours on average between the two sample means, what would your conclusion be? How would your answer change if the difference in sample means was only 3? Why?






*Now, let's try to find the value of the difference in means* that corresponds to 5% of the simulated values in the upper tail of the null distribution.

```{r, adding rejection region to the plot}
# --- Step 3: Calculate the Critical Value (95th Percentile) ---
critical_value <- quantile(null_differences, 0.95)

# Print the value to the console so you can see it
#cat("The Critical Value (Top 5% cutoff) is:", round(critical_value, 2), "\n")

# --- Step 4: Visualize ---
null_data <- data.frame(diff_means = null_differences)

ggplot(null_data, aes(x = diff_means)) +
  # 1. Base Histogram (Gray)
  geom_histogram(binwidth = 0.5, fill = "lightgray", color = "black") +
  
  # 2. Highlighted Histogram (Red) - Only data >= critical value
  geom_histogram(data = subset(null_data, diff_means >= critical_value),
                 binwidth = 0.5, fill = "red", color = "black") +
  
  # 3. Vertical Line at Critical Value
  geom_vline(xintercept = critical_value, color = "darkred", linetype = "dashed", size = 1) +
  
  # 4. Label for the Top 5%
  annotate("text", x = critical_value, y = Inf, label = "Top 5% of the\ndistribution", 
           vjust = 1.5, hjust = -0.1, color = "darkred", fontface = "bold") +
  
  # 5. Labs and Theme
  labs(
    title = "Null Distribution of Difference in Sample Means (1000 simulations)",
    subtitle = "Assuming Null Hypothesis is True (Mean Diff = 0)",
    x = "Difference in Means (Women - Men)",
    y = "Count"
  ) +
  theme_minimal()
```

> 11. Based on the plot above, how big of a difference in sample means do you need in order to have 5% of the data in the upper tail of the distribution? ( In other words, where does the red dotted line seem to be located on x axis?)





_You should find a difference of roughly 3.7. The fact that a difference of 3.7 in the two sample means is what separates the top 5% of the data from the rest of the null distribution means that, at a significance level of 0.05, differences in sample means of more than 3.7 will be statistically significant (p-value $\le 0.05$), whereas differences of less than 3.7 between the two sample means will not be statistically significant (p-value \> 0.05). In other words, the sample average number of hours/week women spend parenting would need to be at least 3.7 hours larger than the sample average for mean for us to be convinced to reject the null hypothesis._ The region difference in group means greater or equal to 3.7 is called the **rejection region*.

## Definition

The *rejection region* tells us how different the sample statistic (e.g., difference in group means) need to be in order for us to reject the null hypothesis at a specified level of significance.

So, what does the *rejection region* tell us about how big our sample sizes need to be? In order to answer this question, we have to make an assumption about what the true difference in average hours of parent is. For example, what if, in reality, the average number of hours (per week) women spend parenting is, say, 8 hours more (a full “work day’s” worth) than that for men?

> 12. Now, let's change the population mean for population of men to be 7 hours (this is if men parented 8 hours less than women and women were at 15 hours per week). Take a single sample of size 10 men and 10 women. How different are the group means? (Note: we are no longer assuming that the null hypothesis is true and instead operate under the assumption that the alternative hypothesis is true)

```{r, simulates single random sample for women and men respectively with updated means}
# --- Parameters ---
n_per_group <- 10
mean_women <- 15 
mean_men <- 7
sd_pop <- 5
num_simulations <- 1

# --- Simulation Function ---
simulate_diff_means <- function(n, mu1, mu2, sd) {
  sample1 <- rnorm(n, mean = mu1, sd = sd)
  sample2 <- rnorm(n, mean = mu2, sd = sd)
  return(mean(sample1) - mean(sample2))
}

# --- Step 1: Perform one simulation (like "Draw samples" in applet) ---
set.seed(546) # Reproducibility
sample_women <- rnorm(n_per_group, mean = mean_women, sd = sd_pop)
sample_men <- rnorm(n_per_group, mean = mean_men, sd = sd_pop)

cat("--- Single Sample Summary ---\n")
cat("Women Mean:", round(mean(sample_women), 2), " SD:", round(sd(sample_women), 2), "\n")
cat("Men Mean:  ", round(mean(sample_men), 2), " SD:", round(sd(sample_men), 2), "\n")
cat("Difference in Means:", round(mean(sample_women) - mean(sample_men), 2), "\n")
```




> 13. Now, let's take 1000 samples. What is the mean of the difference in group means across the 1000 samples? What is the SD? How do these values compare to the one’s we obtained earlier when men and women were parenting the same amount on average? Why do these new values make sense?

```{r, simulates 1000 difference in means for women and men respectively with updated population means}
# --- Parameters ---
n_per_group <- 10
mean_women <- 15 
mean_men <- 7
sd_pop <- 5
num_simulations <- 1000

# --- Simulation Function ---
simulate_diff_means <- function(n, mu1, mu2, sd) {
  sample1 <- rnorm(n, mean = mu1, sd = sd)
  sample2 <- rnorm(n, mean = mu2, sd = sd)
  return(mean(sample1) - mean(sample2))
}

# --- Step 1: Perform simulation ---
set.seed(123) 
sample_women <- rnorm(n_per_group, mean = mean_women, sd = sd_pop)
sample_men <- rnorm(n_per_group, mean = mean_men, sd = sd_pop)

# --- Step 2: Perform 1000 Simulations (Null Distribution) ---
alternative_differences <- replicate(num_simulations, simulate_diff_means(n_per_group, mean_women, mean_men, sd_pop))

# --- Calculate Mean and Standard Deviation of the Null Distribution ---
mean_alternative_diffs <- mean(alternative_differences)
sd_alternative_diffs <- sd(alternative_differences)

# --- Step 3: Visualize ---
alternative_data <- data.frame(diff_means = alternative_differences)

ggplot(alternative_data, aes(x = diff_means)) +
  # 1. Base Histogram (Gray)
  geom_histogram(binwidth = 0.5, fill = "lightgray", color = "black") +
  
  # Add a vertical line for the mean of the null distribution
  geom_vline(xintercept = mean_alternative_diffs, color = "blue", linetype = "dashed", size = 1) +
  
  # Add annotations for mean and standard deviation
  # THIS annotate CALL IS COMPLETE AND ENDS WITH A PARENTHESIS ')'
  annotate("text", x = mean_alternative_diffs, y = Inf,
           label = paste("Mean =", round(mean_alternative_diffs, 2)),
           vjust = 1, hjust = -1.5, color = "blue", fontface = "bold") +
  annotate("text", x = sd_alternative_diffs, y = Inf,
           label = paste("SD =", round(sd_alternative_diffs, 2)),
           vjust = 2.5, hjust =-5.6, color = "blue", fontface = "bold")+
    # 5. Labs and Theme
  labs(
    title = "Alternative Distribution of Difference in Sample Means (1000 simulations)",
    subtitle = "Assuming Alternative Hypothesis is True (Mean Diff > 0)",
    x = "Difference in Means (Women - Men)",
    y = "Count"
  ) +
  theme_minimal()
```

> 14. Use the output below to answer this question:On this new distribution, how often did you get values greater than 3.7? Interpret this percentage.

```{r, adding power to the alternative distribution plot}

# --- NEW: Define the fixed highlight value ---
highlight_threshold <- 3.7 # The value at which to start highlighting
proportion_highlighted <- sum(alternative_differences >= highlight_threshold) / length(alternative_differences)

# --- Step 3: Visualize the Null Distribution with Highlighted Tail (at 3.7 and above) ---
alternative_data <- data.frame(diff_means = alternative_differences)

ggplot(alternative_data, aes(x = diff_means)) +
  # 1. Base Histogram (Gray)
  geom_histogram(binwidth = 0.5, fill = "lightgray", color = "black") +
  
  # 2. Highlighted Histogram (Green) - Only data >= highlight_threshold
  geom_histogram(data = subset(alternative_data, diff_means >= highlight_threshold),
                 binwidth = 0.5, fill = "seagreen", color = "black") + # Changed fill to 'seagreen'
  
  # 3. Vertical Line at the Highlight Threshold
  geom_vline(xintercept = highlight_threshold, color = "darkgreen", linetype = "dashed", size = 1) + # Changed color
  
  # 4. Label for the Highlighted Area 
  annotate("text", x = highlight_threshold, y = Inf,
           label = paste("Difference in means ", highlight_threshold), # Updated label text
           vjust = 1.5, hjust = -0.1) +
   annotate("text", x = proportion_highlighted, y = Inf,
           label = paste("Power = ", proportion_highlighted), # Updated label text
           vjust = 3, hjust = -1.25, color = "darkgreen",, fontface = "bold" ) +
    # 5. Labs and Theme
  labs(
    title = "Alternative Distribution of Difference in Sample Means (1000 simulations)",
    subtitle = paste("Probability of rejecting the null when alternative is true - Statistical Power"),
    x = "Difference in Means (Women - Men)",
    y = "Count"
  ) +
  theme_minimal() +
  # Expand Y axis top margin for labels
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
  
```








*About 97% of the time, we get difference in means that are greater than 3.7. This means that if women, on average, parent 8 more hours per week than men and we repeatedly randomly sample 10 men and 10 women from the population, about 97% of the time we would get the sample mean for women that is 3.7 hours or higher than the sample mean for men. The value you’ve estimated is the statistical power*.

## Definition

The *statistical power* of a study is the *probability that the researchers find evidence against the null hypothesis and in favor of the alternative hypothesis when the alternative hypothesis is true*.

To see that this is what we just did remember that we: (a) first found that we will reject the null hypothesis and find evidence for the alternative hypothesis if the difference in group means is more than 3.7; and then (b) found that differences larger than 3.7 will happen approximately 97% of the time if the average parenting time for women is 15 hours/week and only 7 hours/week for men.

> 15. What is your estimate of the Type II error rate in this study? What would a Type II error mean in this context?




## Factors That Impact Power

As we saw in Example 1.6, typically we like to have the power of the study be approximately 80% to consider the study to be “well-powered” and not more than necessary. Thus, having 10 people in each group may be more than we need. There are numerous factors which effect the power of a study. We will explore them now to help us determine the sample size we should have in this study.

## How can we do what we did with simulations with built-in function in R?

```{r, built-in R function to calculate statistical power}
library(pwr)

# --- Scenario Parameters ---
n_women <- 10
n_men <- 10
mean_women <- 15
mean_men <- 7
sd_pop <- 5
alpha <- 0.05 # Significance level

# --- Calculate Effect Size ---
effect_size <- abs(mean_women - mean_men) / sd_pop
cat("Calculated effect size:", effect_size, "\n\n")

# --- Calculate Statistical Power using pwr.t.test ---
power_result <- pwr.t.test(
  n = n_women,       # Sample size per group (assumes equal n)
  d = effect_size,      # effect size
  sig.level = alpha, # Significance level
  type = "two.sample", # Two independent samples
  alternative = "greater" # One-sided test (women mean > men mean)
)

print(power_result)
```

## Sample Size

> 16. What would happen if we sample 5 men and 5 women? Estimate the power of the study to detect a difference of 8 hours per week. Provide an interpretation of power in this context.

```{r, built-in R function to calculate statistical power for difference of 8 and smaller samples}
library(pwr)

# --- Scenario Parameters ---
n_women <- 5
n_men <- 5
mean_women <- 15
mean_men <- 7
sd_pop <- 5
alpha <- 0.05 # Significance level

# --- Calculate Effect Size ---
effect_size <- abs(mean_women - mean_men) / sd_pop
cat("Calculated effect size:", effect_size, "\n\n")

# --- Calculate Statistical Power using pwr.t.test ---
power_result <- pwr.t.test(
  n = n_women,       # Sample size per group (assumes equal n)
  d = effect_size,      # effect size
  sig.level = alpha, # Significance level
  type = "two.sample", # Two independent samples
  alternative = "greater" # One-sided test (women mean > men mean)
)

print(power_result)
```




> 17. What is the relationship between power and sample size? Why does this make intuitive sense?




## Size of the Actual Difference/How Wrong the Null Hypothesis Is

_What if we decided that a difference of eight hours per week was large but were curious about smaller differences? What if the difference was only 4 hours per week (1/2 a work day)? Start by sampling 10 men and 10 women again._

```{r, built-in R function to calculate statistical power for difference of 4 and original samples}
library(pwr)

# --- Scenario Parameters ---
n_women <- 10
n_men <- 10
mean_women <- 15
mean_men <- 11
sd_pop <- 5
alpha <- 0.05 # Significance level

# --- Calculate Effect Size ---
effect_size <- abs(mean_women - mean_men) / sd_pop
cat("Calculated effect size:", effect_size, "\n\n")

# --- Calculate Statistical Power using pwr.t.test ---
power_result <- pwr.t.test(
  n = n_women,       # Sample size per group (assumes equal n)
  d = effect_size,      # effect size
  sig.level = alpha, # Significance level
  type = "two.sample", # Two independent samples
  alternative = "greater" # One-sided test (women mean > men mean)
)

print(power_result)
```

> 19. What is the power if the actual difference in the population means is 4 hours per week?

> 20. How does the power change as the difference in group means changes (or increases/deceases)?



_As we saw earlier in this chapter, the F-statistic is a standardized statistic with a known theoretical distribution. Instead of comparing the difference in means, let’s now use the F-statistic to estimate the power._

> 21. Use the R code below to calculate the power of the test using F distribution. We are interested in estimating the power of detecting a difference of 8 hours in the population means with the F-statistic. How does this power estimate compare to the one you found in #14 when using the difference in means? Why is it a bit different?

```{r, power calculations using F distribution for origninal data}
library(pwr)

# --- Scenario Parameters ---
n_women <- 10
n_men <- 10
mean_women <- 15
mean_men <- 7
sd_pop <- 5 # Common standard deviation for all groups in ANOVA context
alpha <- 0.05 # Significance level

# --- Parameters for pwr.anova.test ---
k_groups <- 2 # Number of groups (women, men)
# For pwr.anova.test, 'n' is the sample size *per group*.
# Assuming equal group sizes, n_per_group = n_women (or n_men).
n_per_group <- n_women

# --- Calculate Effect Size for ANOVA ---
grand_mean <- (mean_women + mean_men) / k_groups

# Sum of squared deviations of group means from the grand mean
sum_sq_dev_means <- (mean_women - grand_mean)^2 + (mean_men - grand_mean)^2

# Calculate effect size
f_effect_size <- sqrt(sum_sq_dev_means / (k_groups * sd_pop^2))

cat("Calculated Grand Mean:", grand_mean, "\n")
cat("Calculated Cohen's f for ANOVA:", f_effect_size, "\n")


# --- Calculate Statistical Power using pwr.anova.test ---
power_result_anova <- pwr.anova.test(
  k = k_groups,          # Number of groups
  n = n_per_group,       # Sample size per group
  f = f_effect_size,     # Cohen's f effect size
  sig.level = alpha      # Significance level
)

# Print the power analysis result
print(power_result_anova)

```


## Significance Level

What if we were more concerned about Type I errors and so wanted to _set the Type I error rate to 1%_? If we sampled 10 men and 10 women, the SD is 5 hours in each group, and we are back to using the difference in means as our statistic and a one-sided test. 

> 24. What is the power to detect a difference of 8 hours per week (using the difference in means)?

```{r, built-in R function to calculate statistical power for difference of 8 and smaller probability of Type I error}
library(pwr)

# --- Scenario Parameters ---
n_women <- 10
n_men <- 10
mean_women <- 15
mean_men <- 7
sd_pop <- 5
alpha <- 0.01 # updated significance level=P(Type I Error)

# --- Calculate Effect Size ---
effect_size <- abs(mean_women - mean_men) / sd_pop
cat("Calculated effect size:", effect_size, "\n\n")

# --- Calculate Statistical Power using pwr.t.test ---
power_result <- pwr.t.test(
  n = n_women,       # Sample size per group (assumes equal n)
  d = effect_size,      # effect size
  sig.level = alpha, # Significance level
  type = "two.sample", # Two independent samples
  alternative = "greater" # One-sided test (women mean > men mean)
)

print(power_result)
```

> 25. How does the power change when the Type I error rate decreases?



## Population Standard Deviation

To this point, we’ve been assuming that the standard deviation within each group is 5 hours. But what if it isn’t?

> 26. Let's calculate the power of detecting a difference of 8 hours a week if we sample 10 men and 10 women, but the SD is 2.5 hours in each population. How is power affected by changes in the population SDs?

```{r, built-in R function to calculate statistical power for difference of 8 and smaller SDs}
library(pwr)

# --- Scenario Parameters ---
n_women <- 10
n_men <- 10
mean_women <- 15
mean_men <- 7
sd_pop <- 2.5
alpha <- 0.01 # updated significance level=P(Type I Error)

# --- Calculate Effect Size ---
effect_size <- abs(mean_women - mean_men) / sd_pop
cat("Calculated effect size:", effect_size, "\n\n")

# --- Calculate Statistical Power using pwr.t.test ---
power_result <- pwr.t.test(
  n = n_women,       # Sample size per group (assumes equal n)
  d = effect_size,      # effect size
  sig.level = alpha, # Significance level
  type = "two.sample", # Two independent samples
  alternative = "greater" # One-sided test (women mean > men mean)
)

print(power_result)
```


> 27. Consider this: Suppose that when recruiting the participants for the study, the researcher selected heterosexual couples with children, and surveyed both the male and female parents in the pair, comparing the observations within each pair to estimate the male/female difference. Do you think that this design will have an effect on the statistical power of the test? How and why? (Hint: do you think the responses of parents in the same pair will be independent?)






_The study design outlined in #28 uses *matched pairs*. In the next chapter, we’ll be exploring how statistical power is affected by whether the study is a matched pairs study or not and other things you can do to design your study to impact the power of the study._

**Summary**

In this section, you examined the factors that impact the power of a study, the probability that the study will find a statistically significant p-value when conducted in a situation where the null hypothesis (no association) is false. These factors included:

-   the level of significance,
-   the sample size,
-   the number of groups, and
-   the strength of the actual association.

You should be able to describe how increasing each of these individually lowers or raises the power of a study. The primary use of a power calculation is in designing a study, to determine the sample size that is necessary to have a high enough probability of detecting the type of association that you are interested in.
