```r
---
title: "Math 248: Exploration 2.1"
author: "Your name here"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
```

## Exploration 2.1 Chip Melting Times

Industrial-scale baking requires a precise understanding of how recipe ingredients behave when baked. For example, bakers want to know and understand how different types of baking chips respond to different oven temperatures. In this study you will compare the melting times of chocolate chips and butterscotch chips. You are going to be the participants, and the response variable will be the amount of time, from placing a chip on your tongue and holding it steadily to the roof of your mouth, until the chip is completely melted. Of course, there are more precise ways of measuring melting times, but this approach is tastier!

1. **Study Protocol:**

    *   **Design:** A matched-pairs design will be used, where each participant will test both chocolate and butterscotch chips. The order of the chips will be randomized to minimize order effects.  This is a repeated measures design.
    *   **Random Assignment:**  Participants will be randomly assigned to either the chocolate-first or butterscotch-first group.  This randomization helps to control for any potential time-related biases.
    *   **Inclusion Criteria:**  Participants must be able to distinguish the taste and texture of the chips and should not have any known allergies or sensitivities to chocolate or butterscotch.  Participants should also be comfortable with the unusual measurement method.  A sample size calculation should be undertaken to ensure adequate statistical power.

2. **Sources of Variation Diagram:**

| Observed variation in | Sources of explained variation | Sources of unexplained variation |
|-----------------------|--------------------------------|---------------------------------|
| Melting Time          | Chip Type (Chocolate/Butterscotch) | Individual Differences (Taste sensitivity, saliva production, oral temperature) |
|                       |                                | Measurement Error (Subjective method) |
|                       | Person                           | Environmental Factors (Room temperature, time of day)|

3. **Addressing Unexplained Variation:**

The sources of unexplained variation (individual differences, measurement error, environmental factors) share the commonality of being potentially uncontrollable or difficult to measure precisely. To improve the study design, the following can be considered:

*   **Standardization:**  Establish clear guidelines for the melting time measurement procedure (e.g., room temperature, amount of chip, holding technique).  Establish inclusion/exclusion criteria for participants that minimize expected individual differences in response.
*   **Blinding:** If possible, blind participants as to the type of chip they are currently testing to reduce bias.
*   **Replication:** Increase the sample size (number of participants) to reduce the influence of random error.
*   **More Objective Measurement:** Consider using a more objective method to measure melting time.


4. **Repeated Measures Design Feasibility:**

A repeated measures design is feasible and highly beneficial in this study. Each participant serves as their own control, reducing the impact of individual differences (a significant source of variation). The benefits include:

*   Increased statistical power due to reduced variability.
*   The ability to directly compare melting times within individuals rather than across different individuals.
*   Fewer participants are needed to reach the same level of statistical power.

5. **Random Assignment in Matched Pairs:**

Random assignment is still crucial in a matched pairs design, even though participants receive both treatments. Randomizing the order of the chips (chocolate first or butterscotch first) is essential.  Without it, any systematic difference (e.g., faster melting of the first chip) could confound the results.


6. **Two-Sample t-test Appropriateness:**

A two-sample t-test is *not* appropriate for this study design.  The two-sample t-test assumes independence between observations. In a repeated measures design, the melting times for chocolate and butterscotch are not independent within each participant, so the assumptions of a two-sample t-test are violated.  The correct analysis would be a paired t-test or repeated measures ANOVA.


7. **R Analysis (using MeltingTimes.txt):**

```{r}
# Download and read the data
data_url <- "https://www.isi-stats.com/isi2/data/MeltingTimes.txt"
melt_data <- read.table(file = data_url, header = TRUE)

# Rename columns for clarity (assuming the order is correct in the file)
names(melt_data) <- c("Person", "ChipType", "MeltingTime")

# Convert ChipType to factor
melt_data$ChipType <- as.factor(melt_data$ChipType)

# Overall mean melting time
overall_mean <- mean(melt_data$MeltingTime)
cat("Overall mean melting time:", overall_mean, "\n")

# Total sum of squares
SSTotal <- sum((melt_data$MeltingTime - overall_mean)^2)
cat("SSTotal:", SSTotal, "\n")

# Mean melting times by chip type
mean_melting_times <- aggregate(MeltingTime ~ ChipType, data = melt_data, mean)
cat("Mean melting time of chocolate:", mean_melting_times$MeltingTime[1], "\n")
cat("Mean melting time of butterscotch:", mean_melting_times$MeltingTime[2], "\n")

# Effect of chocolate and butterscotch (assuming effects are defined as the difference from the overall mean)
effect_chocolate <- mean_melting_times$MeltingTime[1] - overall_mean
effect_butterscotch <- mean_melting_times$MeltingTime[2] - overall_mean
cat("Effect of chocolate:", effect_chocolate, "\n")
cat("Effect of butterscotch:", effect_butterscotch, "\n")


#Paired t-test
t_test_result<-t.test(MeltingTime~ChipType, data=melt_data, paired=TRUE)
t_test_result

```

Remember to install the `tidyverse` package if you haven't already (`install.packages("tidyverse")`).  The above code performs a paired t-test, which is appropriate for this repeated measures design with two treatment groups.  For more than two chip types, you would need a repeated measures ANOVA.  The code currently only calculates effects as deviations from the overall mean; a more complete analysis would involve ANOVA to determine statistical significance.

