---
title: "Math 248: Exploration 1.1"
author: "Your name here "
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
library(ggplot2)
library(dplyr)
library(mosaic)
```


# Exploration 1.1: Sources of Variation (Memorizing Letters)

You and your classmates will each be asked to study a sequence of letters for 20 seconds and then write down as many letters as you can remember, in order. Your score will be the number of letters that you remember correctly before your first error of any kind. You will also be asked to report your hours of sleep last night and whether or not you have consumed any caffeine today. After collecting the data, your instructor will tell you more information about the study protocol.


>1. What was the research question for this study?


>2. Identify the response variable. Is this variable quantitative or categorical? 


>3. Was this an observational study or experiment? Justify your choice. 


## Definition: The *explanatory variable* deliberately manipulated in an experiment is often called a treatment variable or factor. For categorical explanatory variables, the different categories of the treatment variable are often called *levels*. In experiments, the objects (or “subjects”) that we are measuring are often called *experimental units* rather than observational units. The conditions we impose on the experimental units (here the levels of the treatment variable) are also called treatments. Each experimental unit is assigned to one treatment. 



>4. Identify the experimental units in this study. How many are there?

>5. Identify the treatment variable and its level. 

>6. If you were planning this study, how would you determine who gets which treatment? What would you try to accomplish?


## Key Idea: Random Assignment

The goal of random assignment is to reduce the chances of there being any confounding variables in the study. By creating groups that are expected to be similar with respect to variables (other than the treatment variable of interest) that may impact the response, random assignment attempts to eliminate confounding. A key consequence of not having variables confounded with the treatment variable in a randomized experiment is the potential to draw cause-and-effect conclusions between the treatment variable and the response variable. 

>7. Identify any other precautions taken in this study to try to make sure the two treatment groups were equally balanced,



>8. Was this a double-blind or a single-blind study? Explain. What other, if any, appropriate precautions were taken in carrying out this study?


## Definition: Blind Studies 

A study is double-blind if
(i) the subjects do not know which treatment condition they are in, and 
(ii) the person evaluating the response variable does not know which treatment condition the subject is in. 

If only one of the above two criteria are true for a study, then the study is said to be single-blind


>9. Were you and your classmates randomly selected to participate in this study? Do you think you are representative of a larger population? What would you define that population to be?



>10. Are there any individuals that you believe we should exclude from participating in this study? Explain.

## Definition: *Inclusion criteria* are the set of characteristics that individuals must have in order to participate in a study


# Single Sample Mean:

First we will look at the distribution of the data without any explanatory variable, just looking at distribution of scores. 


```{r, EVAL = FALSE}
#load in your class data: 

class_data <- read.csv("your file path here")

#Sample Distribution
ggplot(class_data, aes(x=scores))+   # ggplot() starts a plot, aes() tells it what to use for x-axis
  geom_histogram(fill="skyblue", color = "black")+ # geom_histogram() makes a histogram
  theme_classic()+                   # changes the plot style
  labs(                              # labs() lets you set labels
    title = "Sample Distribution of Scores",
    x = "Scores",
    y = "Frequency")


```



## Key Idea: Residuals are a measure of how far each observation is away from the mean, Lower residual value indicates sample value close to the mean, 



```{r, EVAL = FALSE}
#Obtain Residuals
class_data <- class_data %>% 
  mutate(residuals = scores - mean(scores))   # mutate() creates a new column; here each score minus the overall mean


#Plot Residuals
ggplot(class_data,aes(x=residuals))+
  geom_histogram(fill = "skyblue", color = "black") + # same histogram code, now just with residuals instead of scores. 
  labs(
    title = "Histogram of Residuals (Single Mean Model)",
    x = "Residuals",
    y = "Frequency"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # adds a vertical reference line at 0 
  theme_minimal()
```


>11. Comment on the shape, spread, center, and outliers for the distribution of scores. Be sure to use numerical summaries of center and spread. 

```{r, EVAL = FALSE}
#hint : use favstats from the mosaic package
*your code here*
```


>12. Specify a statistical model for predicting future results using the overall mean score. 



# Seperate by treatment groups:

We add in the factor to compare the two sample means by treatment group by grouping and computing two separate means. 

```{r, EVAL= FALSE}
#Mutating the data to add separate means by treatment as well as separate means residuals. 

class_data <- class_data%>% 
  group_by(trt)%>%                                # group_by Tells R to do calculations separately by treatment group
  mutate(trt_residuals = scores- mean(scores))%>% 
                                                  # within each trt group, calculates residual for each observation
  ungroup()


# Distribution of Response

ggplot(class_data, aes(x=scores, fill=trt))+      # fill=trt colors bars by treatment group
  geom_histogram()+
  facet_wrap(~trt,ncol=1)+                        # facet_wrap() makes one histogram per treatment group (stacked vertically here)
  theme_classic()+
  labs(
    title = "Sample Distribution of Scores (Seperate-Means Model)",
    x = "Scores",
    y = "Frequency")
```


```{r}
#Plot Residuals

ggplot(class_data,aes(x=trt_residuals,fill=trt))+ # plotting residuals based on seperate group means
  geom_histogram() +
  labs(
    title = "Histogram of Residuals (Seperate-Means Model)",
    x = "Residuals",
    y = "Frequency"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_classic()
```


```{r}
# Gives numerical summaries of separate means (still using favstats)
favstats(~scores | trt, data = class_data)

```

>13. Comment on the shape, spread, and centers of the two distributions. Use numerical summaries and comment any differences that you see. 


>14. Write out a statistical model for predicting outcomes depending on which treatment condition someone is assigned to, using the treatment-specific mean scores. This could be called the “letter grouping model” or the “separate-means model.” 


>15. Does knowing which treatment group each person was assigned to explain all of the variation in the responses? How do you decide?


>16. Summarize your “letter-grouping model” by listing all sources of variation. Brainstorm some possible sources of unexplained variation.


Inclusion Criteria: 
 - Sources of explained variation: 
 - Sources of unexplained variation: 


Design: 
 - Sources of explained variation: 
 - Sources of unexplained variation:
 

# Confounding Variables: 

One potential source of variation in how many letters someone can memorize is the amount of sleep he or she got the night before. Ideally, the study protocol has balanced this variable between the two treatment groups (i.e., there is a similar mix of those with lots and little sleep in both groups), but let’s check. Remember, to be a confounding variable, the variable must explain variation in the observed memory scores AND it must be associated with the explanatory variable, type of letter grouping. 


>17. Change the response variable to sleep. Does amount of sleep appear to be a confounding variable in this study? How do you decide? 


```{r, EVAL=FALSE}
#Plotting Sleep Distribution

ggplot(class_data, aes(x= #your code here (hint: What variable are we looking at?)
                        , fill=trt))+
  geom_histogram()+
  facet_wrap(~trt,ncol=1)+
  theme_classic()+
  labs(
    title = "Sample Distribution of Sleep",
    x = "Sleep (hours)",
    y = "Frequency")

```



>18. Change the response variable to caffine. Does amount of caffine appear to be a confounding variable in this study? How do you decide?


```{r, EVAL=FALSE}
#Plotting Caffeine Distribution


ggplot(class_data, aes(x= # your code here (hint: What variable are we looking at?)
                        , fill=trt))+
  geom_histogram(alpha=0.5)+
  facet_wrap(~trt,ncol=1)+
  theme_classic()+
  labs(
    title = "Sample Distribution of Caffeine",
    x = "Caffine (number of drinks)",
    y = "Frequency")

```



>19. Could there be another explanation, apart from the grouping of the letters, that could explain the difference in the group means that we found? Explain.



## Key Idea: 

Random assignment is a critical component of a well-designed experiment that allows us to potentially draw cause-and-effect conclusions. How the experimental units are selected for the study (e.g., inclusion criteria, random sampling) is a key component of how broadly we can generalize the results.


Hopefully you found that the conditional distributions of the amount of sleep for the two treatment groups and the conditional distributional distributions of caffeine for the two treatment groups were pretty similar, thus preventing amount of sleep and caffeine from being confounding variables. With random assignment, we will often trust that this is true for all other potential confounding variables as well.



>20. In order to make predictions about how many letters students will be able to memorize, do you recommend the single-mean model or the letter-grouping model? What population are you willing to generalize these observations to? Are you willing to draw a cause-and-effect conclusion between the type of sequences and the ability to memorize more letters on average in this population? Explain. 



>21. Remember to take "EVAL=FALSE" out of all R chunks. 
