---
title: "Math 248: Exploration 2.1"
author: "Your name here"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
library(ggplot2)
library(dplyr)
library(ImSIwithR)
library(readr)
```
  
```{r}
#load in data
data(MeltingTimes)
```

# Exploration 2.1 - Chip Melting Times

Industrial-scale baking requires a precise understanding of how recipe ingredients behave when baked. For example, bakers want to know and understand how different types of baking chips respond to different oven temperatures. In this study you will compare the melting times of chocolate chips and butterscotch chips. You are going to be the participants, and the response variable will be the amount of time, from placing a chip on your tongue and holding it steadily to the roof of your mouth, until the chip is completely melted. Of course, there are more precise ways of measuring melting times, but this approach is tastier!

> 1.  Outline a study protocol for investigating this question. Specifically, how will you design your study? How will you use random assignment? What about inclusion criteria? (Reminder: Inclusion Criteria is key features of the target population of interest.)



> 2.  Brainstorm a Sources of Variation diagram, including discussion of any “controls” you will use in the study to minimize unexplained variation.


 | Observed variation in | Sources of explained variation | Sources of explained variation |
 |-----------------------|--------------------------------|--------------------------------|
 | Inclusion criteria    |                                |                                |
 |                       |                                |                                |
 | Design                |                                |                                |



> 3.  Reviewing your sources of unexplained variation, do they have anything in common? Based on this commonality, can you suggest a way to improve the design of your study to try to address these sources?



### Definition : Matched Pairs Design

A *matched pairs design* pairs experimental units that are similar to each other and randomly assigns each member within a pair to one of the two treatment groups. A special type of matched pairs design is **repeated measures** where the same individual receives both treatments in a randomly assigned order.

```{r}
ggplot(MeltingTimes, aes(time, fill = chiptype))+
  geom_histogram(color="white")+
  theme_classic()+
  labs(title="Melting Times Distribution by Chiptype")
```


> 4.  Is a repeated measures type of matched pairs design feasible in this study? Describe how it would work and its main benefits.





### Key Idea

By having the same experimental unit (e.g., person) complete both treatments, the variability from experimental unit to experimental unit (e.g., person to person) is held constant and is no longer “unexplained variation.”


> 5.  Does random assignment still play an important role in a matched pairs design? Explain why it matters and what could happen if you didn’t use random assignment.



> 6.  Would a two-sample t-test (or a separate means model) be an appropriate analysis to compare the two melting time means for our matched pairs study design? Explain.



### Key Idea

In a paired design, it is not reasonable to consider the multiple observations on the same individual as independent, so we cannot carry out a procedure (e.g., a two-sample t-test) that assumes we have two independent samples.

To analyze a paired design you can either use a one-sample t-test (and confidence interval) on the differences or a simulation that maintains the paired nature of the study design (e.g., randomizing the order of the two treatments on the same individual). The validity condition of the one-sample t-test is that the distribution of the differences is reasonably symmetric or the sample size is large with no severe skewness or outliers (e.g., larger than 30).

### To work through a simulation-based analysis that simulates random assignment of the treatment within each pair or to conduct a one-sample t-test on the differences (i.e., the paired t-test that you may have seen in your first course), refer to the Homework Exercises. The paired t-test approach only works with two treatments and so we will show you a more general method here. (Data from one of the authors’ classes can be found in MeltingTimes.)



```{r, summary statistics without accounting for person effect}
# Find overall mean 
overall_mean <- mean(MeltingTimes$time) # calls the time column of MeltingTimes and finds the average


# Calculating average melting time per chip type, number of observations, effect, and Sum of Squares for the Chip
summary_table <- MeltingTimes %>%
  group_by(chiptype) %>%
  summarise(
    n = n(),
    mean_melting_time = mean(time)) %>%
  mutate(
    effect = mean_melting_time - overall_mean,
    SS = n * effect^2)%>%
  ungroup() 


# Calculate Sum of Squares totals
SSTotal <- sum((MeltingTimes$time - overall_mean)^2)
SSchip <- sum(summary_table$SS)
SSError <- SSTotal - SSchip


# select elements for a summary table (specifying names and variables).
final_table <- summary_table %>%
  select(
    Chip = chiptype,
    `Group Size (n)` = n,
    `Mean Melting Time` = mean_melting_time,
    Effect = effect,
    `SS (Chip)` = SS
  )


# view the table 
final_table
```



```{r}
# Make a simple sum of squares table
sstable <- 
  tibble(
    "SS Total" = SSTotal,
    "SS Chip" = SSchip,
    "SS Error" = SSError
  )
sstable

```

> 7. Write a couple sentences discussing any key features from the previous tables. 





# Person Effect (SSperson):

The data in `MeltingTimes(Exploration2.1) - Sheet1.csv` file posted on Canvas. Upload the `MeltingTimes(Exploration2.1) - Sheet1.csv` file to R studio.

Then, read this `MeltingTimes(Exploration2.1) - Sheet1.csv` file into R Markdown with the commands included in the R chunk below.

*Hint*: to get the correct path to include between the quotation marks, locate the csv file in the *Files* tab, click on the down arrow to the right of *More(Blue Wheel Icon)* and choose *Copy folder path to Clipboard* from the drop-down menu.

In my case the path copied to Clipboard is *"\~/Documents/Math 248 Materials/Math 248 Resources/Data"*. Paste this path before `MeltingTimes(Exploration2.1) - Sheet1.csv` within the quotation marks in R code chunk below. Thus, *my full path* is :*"\~/Documents/Math 248 Materials/Math 248 Resources/Data/MeltingTimes(Exploration2.1) - Sheet1.csv".*

Use `head` command to make sure that the file is read correctly.

Remove **eval=FALSE** after running the code.



```{r, EVAL = FALSE}
# Read your class data from csv
class_data <- read_csv("~/Documents/Math 248 Materials/Math 248 Resources/Data/MeltingTimes(Exploration2.1) - Sheet1.csv")
head(class_data)
```

```{r}
# Merges your class data with the provided dataset
merged_data <- bind_rows(MeltingTimes,class_data)
```

```{r, summaries and statstics for combined dataset (with your class data)}
# overall mean and SSTotal
overall_mean_merged <- mean(merged_data$time, na.rm = TRUE)
SSTotal_merged <- sum((merged_data$time - overall_mean_merged)^2, na.rm = TRUE)

# By chip effects
chip_stats <- merged_data %>%
  group_by(chiptype) %>%
  summarise(
    n = n(),
    mean_time = mean(time, na.rm = TRUE)
  ) %>%
  mutate(
    effect = mean_time - overall_mean_merged,
    SS_Chip = n * effect^2
  )

SSChip_merged <- sum(chip_stats$SS_Chip)

# 
person_stats <- merged_data %>%
  group_by(student) %>%
  summarise(
    n = n(),
    mean_time = mean(time, na.rm = TRUE)
  ) %>%
  mutate(
    person_effect = mean_time - overall_mean_merged,
    SS_Person = n * person_effect^2
  )
```


### Key idea: 

The following procedure is easily expandable to more than two treatment groups. The goal of this analysis is to take the paired nature of the design into account. That is, we want to account for the person-to-person variation. To do that, we need to determine how much variation in the data is explained by the participants. Then, once we determine the *SSperson*, we can take that variation out of the unexplained *SSError.* To find the *SSperson* we need to know the effect of each person. As with the treatment effects, the person effects can be found by (“group” mean − overall mean).



> 8. Choose one person (you, if you recognize your values in the data table) and calculate the average of that person’s two melting times.

Remove **eval=FALSE** after running the code.

```{r, eval=FALSE}
# Filter for one specific person and calculate their mean
single_person_mean <- merged_data %>%
  filter(student == "Initials") %>%   # Replace "Initials" with your initials or the person you want to select.
  summarise(
    mean_time = mean(time, na.rm = TRUE) #takes the mean time
  )

single_person_mean
```


> 9. How does your person's average melting time compare to the overall mean melting time? This is the “effect” of you (or the person you chose).


```{r}
overall_mean_melting_time <- mean(merged_data$time)
cat("Overall mean melting time:",overall_mean_melting_time,"\n")
cat("Single person mean melting time:", single_person_mean$mean_time)
```



> 10. Find the effect of each person by doing the following: 


```{r}
# Find the difference in mean for each person to the overall mean. (overall mean found in 9)
all_person_means <- merged_data %>%
  group_by(student) %>%
  summarise(
    n=n(),
    mean_time = mean(time, na.rm = TRUE),  #mean for each person
    person_effect = (mean_time-overall_mean_melting_time)  #effect of each person
  )

head(all_person_means)
```




> 11. Explain what it means for a person to have a larger positive effect than another person. Explain what it means for a person to have a larger negative effect.



> 12. Now we find SSperson (sum of squares person), found by squaring each person effect, multiplying by the number of trials per each effect (2), then taking the summation. 


```{r}
# Using R as a calculator
SSperson_merged <- sum(2*(all_person_means$person_effect^2))    #uses the equation for sum of squares. 
cat("Sum of Squares (person effect):",SSperson_merged)
```




> 13. Look at the ANOVA table and recognize any similarities. 

```{r}
#Find the SSError by taking the total - sschip and SSperson 
SSError_merged <- SSTotal_merged - SSChip_merged - SSperson_merged

# Calculate Degrees of Freedom
N <- nrow(merged_data)
k_chips <- n_distinct(merged_data$chiptype)
k_people <- n_distinct(merged_data$student)

df_chip <- k_chips - 1
df_person <- k_people - 1
df_error <- (N - 1) - df_chip - df_person 

# Build the Table with all Sum of Squares, along with degrees of freedom
anova_table <- tibble(
  Source = c("Chip", "Person", "Error", "Total"),
  df = c(df_chip, df_person, df_error, N - 1),
  SS = c(SSChip_merged, SSPerson_merged, SSError_merged, SSTotal_merged),
  MS = c(SSChip_merged/df_chip, SSPerson_merged/df_person, SSError_merged/df_error, NA)
)

print(anova_table)
```


> 14. What proportion of the total variation is explained by chip type? What proportion is explained by person? Interpret. 

```{r}
#makes a table with proportion of total variation. 
variations <- tibble(
  `Prop Explained by Chip` = SSChip_merged / SSTotal_merged,
  `Prop Explained by Person` = SSPerson_merged / SSTotal_merged,
  `Unexplained (Error)` = SSError_merged / SSTotal_merged
)

#shows the table
print(variations)
```


> 15. Now compute an F-statistic for comparing the melting times of the chip types. Is this person-adjusted F-statistic significantly different? What does this mean in the context of the problem? (Hint: chocolate has a higher melting average time.)

```{r}
# Mean Squares used for F statistic
MS_Chip <- SSChip_merged / df_chip
MS_Error <- SSError_merged / df_error


# F_statistic calculated with 
F_stat <- MS_Chip / MS_Error

# Calculate p-value (significance)
p_value <- pf(F_stat, df_chip, df_error, lower.tail = FALSE)

cat("F-Statistic:", F_stat, "\n")
cat("P-value:", p_value, "\n")
```


> 16. Write out a prediction equation for the predicted melting time based on the individual and the type of chip. (Hint: overall mean should also be included)




> 17. Find a prediction for one individual based on your prediction equation. 


Remove **eval=FALSE** after running the code. 


```{r, eval=FALSE}
# Making a prediciton based on your inputs: 

target_student <- "Initials"      # Change to your student code
target_chip <- "chiptype"        # change to the chip type you want to predict ("butterscotch" or "chocolate")


# call for the effect of chiptype: 
chip_effect_val <- chip_stats %>% 
  filter(chiptype == target_chip) %>% 
  pull(effect)



# call for the person effect of your specified student
person_effect_val <- person_stats %>% 
  filter(student == target_student) %>% 
  pull(person_effect)



# calculate prediction using the model equation. 
predicted_time <- overall_mean_merged + chip_effect_val + person_effect_val


#format output 
cat("Base Average:", overall_mean_merged, "\n",
    "+ Chip Effect:", chip_effect_val, "\n",
    "+ Person Effect:", person_effect_val, "\n",
    "----------------------------", "\n",
    "Predicted Time:", predicted_time)
```

> 18. If there were three types of chips being tested, how would your results and analysis change? How would ANOVA calculations change? 


