---
title: "Math 248: Exploration 3.3"
author: "Solution"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE, fig.align = 'center', out.width="80%" )
library(tidyverse)
library(ggplot2)
library(dplyr)
library(mosaic)
library(ImSIwithR)
```

# Exploration 3.3: Hurricane Names

## STEP 1: Ask a research question.

A study by Jung et al. (Proc. Natl. Acad. Sci. USA, 2014) looked at hurricane names and their perceived threat. They thought that if hurricanes had male names they would be perceived as more dangerous than if they had female names. In one part of the study, the participants were given a weather map showing a hurricane moving toward the area where they lived. Participants were told the name of the hurricane and then rated the riskiness of the hurricane on a seven-point scale, with 1 being "not at all" (risky) and 7 being "very risky".

## STEP 2: Design a study and collect data.

Suppose we want to compare reactions to the hurricane names "Arthur" and "Bertha".

> 1.  The researchers decided to block on the participant. Explain what it means to “block on the participant” in this study. Why do you think they chose that blocking variable? How does that change your randomization plan?

Blocking on participant means that each participant will rate both of the hurricanes (in random order). This accounts for the differences of each individual toward hurricane risk.

*Suppose the researchers had two participants in their study. Blocking on participant and randomizing to two treatments is a matched pairs design as discussed in Section 2.1, and a special case of the randomized complete block designs discussed in Section 2.2. Suppose they carried out the matched pairs design for these two participants and obtained the following ratings:*

| Participant | Hurricane name  | Rating (from 1 to 7) |
|-------------|-----------------|----------------------|
| 1           | Arthur (male)   | 3.5                  |
| 1           | Bertha (female) | 4                    |
| 2           | Arthur (male)   | 5                    |
| 2           | Bertha (female) | 4                    |

> 2.  Based on this table, do male names tend to receive higher ratings than female names? Or, can you not judge such a main effect because there appears to be an interaction between participant and rating? How are you deciding?

We cannot judge that main effect because there appears to be an interaction between participant and rating as one participant rated the male hurricane as more dangerous and the other the female hurricane

*We can’t say too much from this study due to the very small sample size. In particular, one participant rated Hurricane Arthur as a larger threat, whereas the other participant picked Bertha as the larger threat. Why is this? Is this a trend? Or just random chance? Is there an interaction between Hurricane Name and Participant?*

> 3.  Fill out the degrees of freedom in the Combined column of the following skeleton ANOVA table. Note: `HG` denotes Hurricane Gender, `prt` denotes Participant, `obs` denotes observation

| Treatment | df        | Design   | df         | Combined               | df    |
|-----------|-----------|----------|------------|------------------------|-------|
|           |           | prt      | 2-1=1      | prt                    | 1     |
| HG        | 2-1=**1** |          |            | HG                     | 1     |
|           |           | obs(prt) | (2-1)\*2=2 | obs(prt)\|HG= residual | 2-1=1 |
| **Total** | 4-1=3     | Total    | 4-1=3      | **Total**              | 4-1=3 |

The corresponding theoretical model is:

`Rating = Hurricane Gender + Participant +  Error`,

*Validity conditions:*

-   Equal SD/Variances

-   Independent observations

-   Normally distributed response variable

*RCBD condition:*

-   No block by treatment interaction

If we were to try to test if there is a violation of the no block by treatment interaction for this model, the Skeleton ANOVA will change to the following. Note: `HG` denotes Hurricane Gender, `prt` denotes Participant, `obs` denotes observation

| Treatment | df | Design | df | Combined | df |
|----|----|----|----|----|----|
|  |  | prt | 2-1=1 | prt | 2-1=1 |
| HG | 2-1=**1** |  |  | HG | 2-1=1 |
| prt\*HG | (2-1)\*(2-1)=1 |  |  | prt\*HG | (2-1)\*(2-1)=1 |
|  |  | obs(prt) | (2-1)\*2=2 | obs(prt)\|HG = residual | **2-2=0** |
| **Total** | 4-1=3 | Total | 4-1=3 | **Total** | 4-1=3 |

## Key Idea: Saturated Model

Because there are only three degrees of freedom total available in the study (4 - 1), all of the degrees of freedom will be used up if an interaction between participant and hurricane name is included. Thus, there will be no way to estimate random error. The interaction effects and the Error are completely confounded. In Chapter 2, we simply assumed there was no interaction between the explanatory variable of interest (here male or female hurricane name) and the blocking variable (participant). But we see some evidence from even these four data values that these two participants are reacting differently to the hurricane names, and we would like to have enough information to test the statistical significance of this interaction.

> 4.  How could you modify the study in order to estimate within-person variation? Would adding more participants to the study help with this issue?

To estimate "within-person" variation, we need more observations for each participant. In other words, we want to give each participant more than two hurricane names. Adding more participants would increase the observation within participant degrees of freedom as well, but at the same time, that will increase the interaction degrees of freedom, and thus, the residual degrees of freedom will still be 0. Let's assume that we increased the number of participants to 10, then the following changes are reflected:

| Treatment | df | Design | df | Combined | df |
|----|----|----|----|----|----|
|  |  | prt | 10-1=9 | prt | 10-1=9 |
| HG | 2-1=**1** |  |  | HG | 2-1=1 |
| prt\*HG | (10-1)\*(2-1)=9 |  |  | prt\*HG | 9 |
|  |  | obs(prt) | (2-1)\*10=10 | obs(prt)\|HG = residual | **10-10=0** |
| **Total** | 20-1=19 | Total | 20-1=19 | **Total** | 20-1=19 |

*In order to estimate within-person variation, simply adding more participants won’t help if we continue to analyze the data as a matched pairs design. This would give us more pairs, but still would not allow us to estimate the “natural variation” in responses by the same people to the same hurricane name.*

## Key Idea: Replication

**Replication** means that a set of experimental or observational study measurements are repeated under similar conditions, allowing for improved estimates of natural (replicate to replicate) variation, which can help improve the evaluation of potential interactions.

In this study, the researchers weren’t that interested in specific hurricane names, but rather the hurricane gender (male or female). Therefore, although it probably wouldn’t make sense to have the same participant rate the same name multiple times, they can rate the same hurricane name gender multiple times. These researchers had participants rate 10 different hurricane names selected from the official 2014 Atlantic Hurricane names, 5 of which were female names (Bertha, Dolly, Fay, Laura, and Hanna) and 5 of which were male names (Arthur, Cristobal, Omar, Kyle, and Marco). Next, we have five repeats for each participant–gender combination, allowing us to estimate the “within participant” variation separately from the between participant and between treatment variation. When you have more than one observation for each block-treatment combination, this is often called a **generalized block design**.

> 5.  Well-designed experiments use random assignment. Where should random assignment be used when conducting this study?

Randomly assign the order that the hurricanes are presented to each participant.

*The data in `hurricanesA` contain the perceived threat level rating given by two participants for each of the 10 hurricanes. Let's load the data.*

```{r, load in data}
library(ImSIwithR)
data(hurricanesA)
hurricanesA<- hurricanesA %>%
  mutate(across(where(is.character), as.factor)) # transforming character variables into factor variables
head(hurricanesA,2)
```

This dataset has 20 observations, thus the Skeleton ANOVA table will look as follows: Note: `HG` denotes Hurricane Gender, `prt` denotes Participant, `obs` denotes observation

| Treatment | df | Design | df | Combined | df |
|----|----|----|----|----|----|
|  |  | prt | 2-1=1 | prt | 2-1=1 |
| HG | 2-1=**1** |  |  | HG | 2-1=1 |
| prt\*HG | (2-1)\*(2-1)=1 |  |  | prt\*HG | (2-1)\*(2-1)=1 |
|  |  | obs(prt) | (10-1)\*2=18 | obs(prt)\|HG = residual | **18-2=16** |
| **Total** | 20-1=19 | Total | 20-1=19 | **Total** | 20-1=19 |

The corresponding theoretical model is:

`Rating = Hurricane Gender + Participant +  Error`,

*Validity conditions:*

-   Equal SD/Variances

-   Independent observations

-   Normally distributed response variable

*RCBD condition (assumed satisfied):*

-   No block by Hurricane Gender interaction

> 6.  Let's create an interaction plot. Describe the nature of the interaction shown in this graph. Do you think it will be statistically significant?

Participant #1 rating the female hurricane as more serious but Participant #346 rating the male hurricane as more serious. This is a pretty large “interaction” and probably statistically significant though we do have a small sample size.

```{r, interaction plot A}
ggplot(hurricanesA,
       aes(x = Participant,
           y = Rating, 
           color = HurrGender, 
           group = HurrGender)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(title = "Interaction Effect of Gender by Participant",
       x = "Participant ID",
       y = "Mean Rating",
       color = "Hurricane Gender") +
  theme_minimal()
```

> 7.  Run the two-way ANOVA model with interaction. Is the interaction significant? Do the degrees of freedom in the output match the degrees of freedom in the Skeleton ANOVA?

With F = 20.61 and pvalue 0.0003, this confirms the interaction seen in the graph was unlikely due to chance alone. The degrees of freedom match the Skeleton ANOVA

```{r, two-way ANOVA with interaction A}
model_interactionA <- aov(Rating ~ Participant*HurrGender, data = hurricanesA)
# View the results
summary(model_interactionA)
```

## When Should We do a RCBD (Randomized Complete Block Design)?

There are cases where generalized block designs are not possible. For example, observational units may get “used up” in the study, eliminating the possibility of replicates. In this case, we can carry out the randomized complete block design from Section 2.2, but we can’t assess the statistical significance of the treatment × block interaction, so we do not include it among the possible sources of variation in the response.

### Key Idea:

In a *randomized complete block design*, with only *one observation in each block* by treatment group combination, *we have to assume there is not a block × treatment interaction*.

## Within Block Factorial Design:

Hurricane names can also be classified by whether they are "old" or "young". In this study, researchers used social security records of names to determine whether a name was an "old" name or a "young" name (based on frequency of use across decades).

## Definition:

A **within-block factorial design** allows you to assess the significance of the factor × factor interaction. Having all treatment combinations in each block gives you enough “replication” to assess the “natural variation” in the responses.

We do not have replication at the treatment level if we focus on the individual hurricane names rather than the gender of the hurricane name. We also need to make sure the random assignment is carried out on the individual "replicates" not only on the blocks.

## STEP 3: Explore the actual data.

In the actual study, the researchers used 346 students from the University of Illinois at Urbana-Champaign participating for course credit (age 19-25 years, 208 females) - each of whom rated each of the 10 hurricane names (5 of each gender). The researchers examined several research questions, but the `hurricanesB` file contains the results for the `hurricane gender` and `hurricane age` variables.

```{r, load hurricanesB data}
data(hurricanesB)
hurricanesB<- hurricanesB %>%
  mutate(across(where(is.character), as.factor)) # transforming character variables into factor variables
head(hurricanesB,2)
```

*Let's focus on hurricane gender and hurricane name ages as two treatment factors in the study. Each subject would rated all four combinations, in random order.*

```{r, interaction plot B}
ggplot(hurricanesB,
       aes(x = HurricaneGender,
           y = Rating,
           color = HurricaneAge,
           group = HurricaneAge)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(title = "Interaction Effect of Hurricane Age by Hurricane Gender",
       x = "Hurricane Gender",
       y = "Mean Rating",
       color = "Hurricane Age") +
  theme_minimal()
```

> 8.  Based on the interaction plot above, does there seem to be an interaction between the Hurricane Gender and Hurricane Age?

Based on the plot alone, comparison of simple effects suggests that there might be a difference and thus potential interaction.

> 9.  Fill out the residual degrees of freedom in a Skeleton ANOVA for updated design. For simplicity I use Gender and Age to represent Hurricane Gender and Hurricane Age in the table.

| Treatment | df | Design | df | Combined | df |
|----|----|----|----|----|----|
|  |  | prt | 346-1=345 | prt | 346-1=345 |
| Gender | 2-1=**1** |  |  | Gender | 2-1=1 |
| Age | 2-1=1 |  |  | Age | 2-1=1 |
| Gender\*Age | (2-1)\*(2-1)=1 |  |  | Gender\*Age | (2-1)\*(2-1)=1 |
|  |  | obs(prt) | (10-1)\*346=3114 | obs(prt)\|G, A, G\*A = residual | 3111 |
| **Total** | 3460-1=3459 | Total | 3460-1=3459 | **Total** | 3460-1=3459 |

The corresponding theoretical model is:

`Rating = Hurricane Gender + Hurricane Age + Hurricane Gender * Hurricane Age + Participant +  Error`,

*Validity conditions:*

-   Equal SD/Variances

-   Independent observations

-   Normally distributed response variable

*RCBD condition (assumed satisfied):*

-   No block by treatment interaction

> 10. Run the two-way ANOVA model with interaction and a blocking factor. Do the degrees of freedom in the output match the degrees of freedom in the Skeleton ANOVA above? Is the interaction significant? Which effects are appropriate to interpret: main effects or simple effects?

```{r, two-way ANOVA with interaction and block Type 3}
# Set global contrasts to 'sum' for Type III SS
options(contrasts = c("contr.sum", "contr.poly"))
# Note: Ensure Participant is treated as a factor if it isn't already
model_interactionB <- lm(Rating ~ factor(ParticipantID) + HurricaneAge * HurricaneGender, data = hurricanesB)# Use lm() instead of aov()

library(car)# Load the package
# 3. Calculate Type III SS
type3_results <- Anova(model_interactionB, type = "III")

# 4. Drop the intercept from the output
# We use [-1, ] to remove the first row
type3_no_intercept <- type3_results[-1, ]

# View the final table
print(type3_no_intercept)
```

### Important Note - Reminder

If your interaction term `(HurricaneAge:HurricaneGender)` is statistically significant (p\<0.05), you *should not* interpret main effects and focus on simple effects only.

If the interaction is significant: It means the effect of `HurricaneGender` changes depending on the `HurricaneAge`. Reporting a *main effect* average can be like saying """The average temperature of a person with one hand in fire and one hand in ice is 'comfortable". It’s mathematically true, but it hides the reality of the situation.

If the interaction is NOT significant: You can safely ignore it and report the *Main Effects* as the primary findings of your study.

### The Estimated Marginal Means (The "Main Effect" Sizes)

Because the interaction is not statistically significant, we will summarize the main effects. We should use Marginal Means (also called Least-Squares Means), which calculate the average effect of one variable while "holding the others constant.

```{r, calculating main effects}
library(emmeans)

gender_means <- emmeans(model_interactionB, ~ HurricaneGender)
cat("MAIN EFFECT: HURRICANE GENDER\n")
cat("Average Rating for each Gender (averaging across Age and Participants)\n")
print(gender_means)

age_means <- emmeans(model_interactionB, ~ HurricaneAge)
cat("MAIN EFFECT: HURRICANE AGE\n")
cat("Average Rating for each Age Group (averaging across Gender and Participants)\n")
print(age_means)

cat("PAIRWISE DIFFERENCES (Direction of the Main Effects)\n")
pairs(gender_means)
pairs(age_means)
```

> 11. Report the sizes of the main effects of Gender and Age. What are the p-values associated with each main effect? Do they match the p-values from the ANOVA output in quesiton 10?

$\text{Main Effect of Hurricane Gender}=-0.165$

$\text{Main Effect of Hurricane Age}=-0.1$

> 12. How much variation in ratings of hurricanes is explained by this model? The R code below outputs Multiple $R^2$ and Adjusted $R^2$. Report both values using appropriate math notation.

The model explains about 20% of the variation in hurricane ratings.

```{r, extract model R-squared}
# Extract Multiple R-squared
r_sq <- summary(model_interactionB)$r.squared
print(r_sq)

# Extract Adjusted R-squared (Recommended for your model)
adj_r_sq <- summary(model_interactionB)$adj.r.squared
print(adj_r_sq)
```

### Why use adjusted $R^2$?

Multiple $R^2$ will always go up as you add more variables, even if they aren't helpful. Because this model includes `ParticipantID` with many levels, we are essentially adding a "dummy variable" for almost every person in your study. Adjusted $R^2$ penalizes the value of $R^2$ for unnecessary variables. We thus should use adjusted $R^2$.

> Examine the residual plots for this model produced by R below. Do you consider the validity conditions for the theory-based F-tests to be met? Include supporting evidence for your conclusions.

We have repeated observations on the individuals, but we included the participant blocking variable in the model, and the order of the treatments was randomized for each subject. We do see some interesting patterns in the residuals on the residual vs fitted plot ,reflecting the blocking on participants and the discrete nature of the scale. The histogram of residuals is approximately symmetric. QQ plot supports the normality of the data since the observations are close to the diagonal line.

```{r, extract residuals and predicted (fitted) values from the model}
plot_data <- data.frame(
  Fitted = predict(model_interactionB),
  Residuals = residuals(model_interactionB)
)
```

```{r, residual vs fitted plot}
ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted Ratings",
       y = "Residuals") +
  theme_minimal()
```

```{r, histogram of the residuals}
ggplot(plot_data, aes(x = Residuals)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Histogram of Residuals",
       x = "Residual Value",
       y = "Frequency") +
  theme_minimal()
```

```{r, normal QQ plot}
ggplot(plot_data, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

> 14. Does this analysis support the researchers’ conjecture that male names would be perceived as more dangerous? (Hint: After adjusting?) Write a short summary of your conclusions, being sure to address significance, estimation, causation, and generalizability.

Yes, after adjusting for participant, there is strong evidence that male hurricanes are perceived as more dangerous. (This study found that this causes problems because people don’t take as many precautionary steps for hurricanes with female names because they aren’t perceived to be as much of a threat.) Because random assignment was employed in the rating order of the hurricanes it can be inferred that the gender of the hurricane name did cause the difference in rating.

> 15. Based on your analysis, is there evidence that blocking on participant was worthwhile? Explain how you are deciding and what this means in context.

Yes, blocking on participant accounted for much of the person to person variation in rating the danger of each hurricane. The significant p-value for participant confirms that. Accounting for these person-to-person differences in how they rate hurricane danger, we were better able to compare the effect of the type of hurricane name on the adjusted ratings.

*Reconsider the hurricane sex/age treatments. The researchers classified the names of the 2014 hurricane names rather than controlling the distribution of this variable. What is a limitation of this approach? (Hint: Consider the number of Old/Male names vs. the number of Young/Female names.) How would you suggest changing the selection of hurricane names used? What is the primary benefit of making this change? The number of results for each treatment was not balanced, ideally there would have been at least 12 hurricanes, such that there would be a balance of 3 hurricanes in each treatment group. A balanced design ensures there is no confounding among the explanatory variables. We will discuss this more in the next section.*

> 16. Explain how to modify the model to consider whether the hurricane gender effect differs across the participants.

| Treatment | df | Design | df | Combined | df |
|----|----|----|----|----|----|
|  |  | prt | 346-1=345 | prt | 346-1=345 |
| Gender | 2-1=**1** |  |  | Gender | 2-1=1 |
| Age | 2-1=1 |  |  | Age | 2-1=1 |
| prt\*Age | (346-1)\*(2-1)=345 |  |  | prt\*Age | (346-1)\*(2-1)=345 |
|  |  | obs(prt) | (10-1)\*346=3114 | obs(prt)\|G, A, p\*A = residual | 3114-347=2767 |
| **Total** | 3460-1=3459 | Total | 3460-1=3459 | **Total** | 3460-1=3459 |

The corresponding theoretical model is:

`Rating = Hurricane Gender + Hurricane Age + Participant +  Error`,

*Validity conditions:*

-   Equal SD/Variances

-   Independent observations

-   Normally distributed response variable

*RCBD condition test is included:*

-   No block by treatment interaction

> 17. Carry out the analysis of the model from the ANOVA table above. Is there a significant interaction between `Hurricane Gender` and `Participant`? What does this mean about the "no block by treatment interaction" validity condition?

Based on the Fstatistic of 1.43 and p-value of \< 0.0001 we have strong evidence that there is a difference in how each participant rates hurricane in regard to what gender the name of the hurricane is. Thus the validity condition of no block by treatment interaction is significant.

```{r, testing gender by participant interaction}
# 1. Set global contrasts to 'sum' for Type III SS (Critical for interactions)
options(contrasts = c("contr.sum", "contr.poly"))

# 2. Update the model to include the Participant x Gender interaction
# Use factor(ParticipantID) * HurricaneGender to get main effects AND the interaction
model_interaction_updated <- lm(Rating ~ HurricaneAge + factor(ParticipantID) * HurricaneGender, 
                                 data = hurricanesB)

library(car)

# 3. Calculate Type III SS
# This will now include a row for "factor(ParticipantID):HurricaneGender"
type3_results_updated <- Anova(model_interaction_updated, type = "III")

# 4. Drop the intercept from the output
type3_no_intercept <- type3_results_updated[-1, ]

# 5. View the final table
print(type3_no_intercept)
```

Since the validity condition of no block by treatment interaction is violated in this case, we would not be able to trust the results of this model. Is there anything we can do? If you take Math 347, you will learn about the type of design structure called the split plot that allows for accounting for block by treatment interactions.

## Key Ideas:

-   Anytime you have a variable that is inherent in your experimental units that you believe is associated with the outcome of interest, it is a good idea to block on it. If you include a block in your design, you must also include it in your analysis. This leads us to a fairly famous set of statistical "rules."

*Block what you can; randomize what you can’t. Replicate as much as possible.*

-   When the validity condition for no block by treatment interaction is violated, transitioning from an RCBD to a split-plot design structure is required. Significant block-by-treatment interaction indicates that treatments react differently across blocks, often due to a block ("whole-plot") factor restricting randomization. The structure changes from a single error term to two, separating main-plot errors from subplot errors.
